---
title: "Box plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Box plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Figures are drawn on the REAL CPB canvas (page = "half" is 2.98 in,
# "full" is 5.96 in; text sizes in theme_cpb() are absolute points, so
# any other canvas distorts the text-to-figure proportion) and only
# *displayed* smaller via out.width, at a consistent 2:1 scale:
# half-page figures at 350px, full-page figures at 700px.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(42)
```

`cpb_box()` draws the CPB distributional figure from **precomputed
quantile columns** (p5/p25/p50/p75/p95; both layers use
`stat = "identity"`, so aggregate your microdata first). The default
style is shown in `vignette("chart-types")`; this vignette covers the
box constructions and combinations.

One data set used throughout -- purchasing power per standard income
group:

```{r data}
groepen <- c("tot 120% wml", "120% wml - mod.", "1 - 1,5x mod.",
             "1,5 - 2x mod.", "2 - 3x mod.", "boven 3x mod.")
raw <- tibble(
  groep      = factor(rep(groepen, each = 400), levels = groepen),
  koopkracht = rnorm(2400, mean = rep(c(-3, -1.5, 0, 1, 2, 3.5), each = 400), sd = 2)
)
kk <- raw |>
  summarise(
    p5  = quantile(koopkracht, 0.05),
    p25 = quantile(koopkracht, 0.25),
    p50 = quantile(koopkracht, 0.50),
    p75 = quantile(koopkracht, 0.75),
    p95 = quantile(koopkracht, 0.95),
    .by = groep
  )
```

# The three box styles

`box_style` selects the construction. `"ggcpb"` (the default, and
the only argument-free call) is the style of the published CPB
distributional figures: capped errorbar whiskers plus an outlined box
with a median line.

```{r box-ggcpb}
cpb_box(kk, x = groep,
  p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "horizontal",
  title    = "Koopkracht per inkomensgroep",
  subtitle = "inkomensgroep",
  ylab     = "% koopkrachtmutatie")
```

`"james"` is the legacy plotter's box: borderless, plain capless
whiskers, a black median line extending past the box, and the median
value printed above it.

```{r box-james}
cpb_box(kk, x = groep,
  p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  box_style   = "james",
  orientation = "horizontal",
  title    = "Koopkracht per inkomensgroep",
  subtitle = "inkomensgroep",
  ylab     = "% koopkrachtmutatie")
```

`"modern"` is the designer variant: light-blue boxes and whiskers, a
thick dark-blue median with the value in bold above it, and the
quartile values printed below the box ends.

```{r box-modern}
cpb_box(kk, x = groep,
  p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  box_style   = "modern",
  orientation = "horizontal",
  width       = 0.35,
  title    = "Koopkracht per inkomensgroep",
  subtitle = "inkomensgroep",
  ylab     = "% koopkrachtmutatie") +
  scale_y_continuous(labels = label_number_nl(accuracy = 0.1))
```

`"james"` and `"modern"` print value labels by default
(`box_labels = FALSE` turns them off, `label_accuracy` controls their
rounding) and draw single-colour boxes: `fill_colour` sets the
colour, and may be a *vector* with one colour per row. A `fill`
mapping is only supported by `"ggcpb"`.

# A fill per year

Map `fill` (with the `"ggcpb"` style) and pass a `position_dodge()`
for grouped boxes -- for example one pair of years per income group:

```{r box-dodged, fig.width = 5.96, fig.height = 2.98, out.width = "700px"}
kk2 <- expand_grid(kk, jaar = factor(c(2026, 2027))) |>
  mutate(across(p5:p95, \(q) q + (jaar == "2027") * 0.8))

cpb_box(kk2, x = groep,
  p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  fill     = jaar,
  position = position_dodge(width = 0.6),
  index    = c(6, 2),
  title    = "Koopkracht per jaar, 2026 en 2027",
  ylab     = "% koopkrachtmutatie") +
  scale_y_continuous(labels = label_number_nl())
```

# Grouped, with a fill per year

The `fill` mapping combines with the vertically grouped layout of
`vignette("layout")`: bold group headings on the category axis, and
within every category a dodged pair of years. `reverse_legend = TRUE`
puts the first year at the bottom of the legend, matching the dodge
order under `coord_flip()`:

```{r box-grouped-fill, fig.height = 4.5}
kk3 <- kk2 |>
  mutate(grp = factor(
    ifelse(groep %in% groepen[1:3], "lagere inkomens", "hogere inkomens"),
    levels = c("lagere inkomens", "hogere inkomens")
  ))

cpb_box(kk3, x = groep,
  p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  fill     = jaar,
  group    = grp,
  position = position_dodge(width = 0.6),
  orientation = "horizontal",
  index    = c(6, 2),
  reverse_legend = TRUE,
  title = "Koopkracht per jaar en inkomensgroep",
  ylab  = "% koopkrachtmutatie")
```

For the single-colour grouped layout (one box per category, one
colour per group, as in the published inkomenseffecten figures), see
`vignette("layout")`.

# Value axis on top

The flagship CPB koopkracht figure draws the value axis along the
*top* of the panel, with the income groups down the side. Pass
`value_axis = "top"` (horizontal boxes only):

```{r box-top}
cpb_box(kk, x = groep,
  p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  box_style    = "modern",
  orientation  = "horizontal",
  value_axis   = "top",
  value_breaks = seq(-6, 6, 2),
  width        = 0.35,
  title    = "Koopkracht per inkomensgroep",
  subtitle = "statisch, verandering in %")
```

Under the hood this sets the value scale to `position = "right"`,
which `coord_flip()` renders along the top edge. Set custom tick
positions through the wrapper's `value_breaks` (as here) rather than
adding a second `scale_y_continuous()`, which would replace the
wrapper's scale.
