---
title: "Chart types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chart types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Figures are drawn on the REAL CPB canvas (page = "half" is 2.98 in,
# "full" is 5.96 in; text sizes in theme_cpb() are absolute points, so
# any other canvas distorts the text-to-figure proportion) and only
# *displayed* smaller via out.width, at a consistent 2:1 scale:
# half-page figures at 350px, full-page figures at 700px.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(42)
```

```{r show-saved-helper, include = FALSE}
# Displays a file save_cpb() already wrote to a temporary path, at a
# stable location knitr manages itself -- so the rendered copy of this
# page (e.g. on GitHub) keeps a working image link rather than one
# pointing at a temp file that no longer exists. Purely a display
# detail for this page, not part of any save_cpb() example above it.
.show_saved <- function(path) {
  stable <- knitr::fig_path(".png")
  dir.create(dirname(stable), recursive = TRUE, showWarnings = FALSE)
  file.copy(path, stable, overwrite = TRUE)
  knitr::include_graphics(stable)
}
```

One section per CPB chart type, each in its *default* house style:
the simulated data it needs and the wrapper call that draws it,
nothing more. All wrappers return a plain `ggplot` object, so
anything here can be extended further with `+`.

Variants and extras live in the other vignettes:
[`vignette("layout")`](layout.html) for facets and grouped category
axes, [`vignette("boxplots")`](boxplots.html) for the box-plot styles
and combinations, [`vignette("annotation")`](annotation.html) for
reference lines and forecast windows,
[`vignette("maps")`](maps.html) for choropleths,
[`vignette("colours")`](colours.html) for picking series colours,
[`vignette("import")`](import.html) for building figures straight
from a CSV, and [`vignette("ggcpb")`](ggcpb.html) for the basic setup
and export.

Two house conventions to know up front:

* **Titles and units.** `title` is the bold heading. The value-axis
  unit goes in `ylab` and is rendered as the italic caption above the
  panel (the CPB "subtitle" position), not as a rotated axis title. A
  titled figure always reserves that caption line, so figures with and
  without a unit align.
* **Zero line.** A solid black zero line is drawn automatically on
  the value axis: always for columns and areas (which are anchored at
  zero), and for lines and boxplots whenever the data spans zero.

# Line charts

A single series needs only `x` and `y`; it is drawn in the CPB
primary blue:

```{r line-single}
bbp <- tibble(
  jaar  = 2015:2027,
  index = 100 + cumsum(rnorm(13, mean = 1.5, sd = 1.2))
)

cpb_line(bbp, x = jaar, y = index,
  title = "Bruto binnenlands product",
  ylab  = "index (2015 = 100)") +
  scale_x_continuous(breaks = seq(2015, 2027, 3), minor_breaks = 2015:2027,
                     guide = guide_axis(minor.ticks = TRUE))
```

Map a column to `colour` for multiple series. The series take the
house colours in their published order -- blue, then magenta -- with
nothing to set. Growth rates span zero, so the black zero line appears
by itself:

```{r line-multi}
groei <- expand_grid(reeks = c("arbeidsproductiviteit", "tfp"),
                     jaar  = 2000:2024) |>
  mutate(waarde = round(rnorm(n(), mean = 1, sd = 1.6), 1))

cpb_line(groei, x = jaar, y = waarde, colour = reeks,
  title = "Productiviteitsgroei",
  ylab  = "%") +
  scale_x_continuous(
    breaks       = seq(2000, 2024, 4),
    minor_breaks = 2000:2024,
    guide        = guide_axis(minor.ticks = TRUE)
  )
```

The added `scale_x_continuous()` shows the usual refinements for a
year axis: labelled breaks every few years plus small minor ticks for
the years in between.

## Markers on the line

A dense time series is drawn as a bare line. When the x axis instead
holds a handful of ordered categories -- age brackets, quintiles --
the house style puts a marker on every observation, so the reader can
see where the measurements actually are. `points = TRUE` adds them in
the line colour, and the legend key becomes a line with a marker on
it:

```{r line-points}
leeftijden <- c("20-30", "30-40", "40-50", "50-60", "60-70", "70-80", ">80")
tarieven <- expand_grid(
  erfenis  = factor(c("10.000 euro", "100.000 euro", "1.000.000 euro"),
                    levels = c("10.000 euro", "100.000 euro", "1.000.000 euro")),
  leeftijd = factor(leeftijden, levels = leeftijden)
) |>
  mutate(tarief = c(14, 12, 13, 10, 8, 10, 12,
                    21, 18, 18, 17, 17, 21, 24,
                    29, 27, 28, 27, 29, 37, 41))

cpb_line(tarieven, x = leeftijd, y = tarief, colour = erfenis,
  points       = TRUE,
  value_limits = c(0, 40),
  title = "Voorkeurstarieven erfbelasting naar leeftijd",
  ylab  = "% gemiddeld voorkeurstarief")
```

`points = TRUE` is the only thing separating this from an ordinary
`cpb_line()` call; `point_size` adjusts the marker if the default
reads too heavy against a thicker line.

# Column charts

`cpb_col()` covers stacked, dodged and horizontal bars. Stacked is
the default `position`:

```{r col-stacked}
sectoren <- c("industrie", "diensten", "landbouw", "overheid")
tw <- expand_grid(jaar   = 2023:2027,
                  sector = factor(sectoren, levels = sectoren)) |>
  mutate(waarde = round(runif(n(), 5, 25), 1))

cpb_col(tw, x = jaar, y = waarde, fill = sector,
  title = "Toegevoegde waarde per sector",
  ylab  = "mld euro")
```

By default the fill legend is reversed (`reverse_legend = TRUE`), so
it reads in the same top-to-bottom order as the stack. For dodged
bars, switch that off so the legend follows the series order:

```{r col-dodged}
scenario <- expand_grid(regio    = c("Noord", "Oost", "Zuid", "West"),
                        scenario = c("basispad", "beleidsvariant")) |>
  mutate(effect = round(runif(n(), -2, 6), 1))

cpb_col(scenario, x = regio, y = effect, fill = scenario,
  position = "dodge",
  reverse_legend = FALSE,
  title = "Effect per regio en scenario",
  ylab  = "% mutatie")
```

## Horizontal bars

`orientation = "horizontal"` flips the chart and moves the gridlines
to the value axis. Here `ylab` labels the *category* axis (still the
caption above the panel) and `xlab` labels the value axis at the
bottom. `pct_axis` formats the value axis with Dutch percentage
labels, `value_limits` fixes its range, and the panel starts exactly
at the zero axis:

```{r col-horizontal}
groepen <- c("tot 120% wml", "120% wml - mod.", "1 - 1,5x mod.",
             "1,5 - 2x mod.", "2 - 3x mod.", "boven 3x mod.")
auto <- tibble(
  # reversed levels put the first-named group at the top after the flip
  inkomensgroep = factor(groepen, levels = rev(groepen)),
  share         = c(62, 35, 26, 19, 14, 13)
)

cpb_col(auto, x = inkomensgroep, y = share,
  orientation  = "horizontal",
  pct_axis     = TRUE,
  value_limits = c(0, 70),
  width        = 0.6,
  title = "Aandeel huishoudens zonder auto naar inkomen",
  ylab  = "inkomensgroep",
  xlab  = "huishoudens zonder auto")
```

A horizontal *dodged* bar needs one extra trick. Under `coord_flip()`
the dodge draws the *last* factor level on top within each group, so
to show (say) 2021 above 2024 you reverse the `jaar` levels and let
`reverse_legend` restore the reading order:

```{r col-horizontal-dodged}
pv <- expand_grid(
  jaar          = factor(c(2021, 2024), levels = c(2024, 2021)),
  inkomensgroep = factor(groepen, levels = rev(groepen))
) |>
  mutate(share = c(10, 14, 19, 22, 26, 33,   # 2021
                   19, 25, 31, 38, 47, 59))  # 2024

cpb_col(pv, x = inkomensgroep, y = share, fill = jaar,
  position     = "dodge",
  orientation  = "horizontal",
  value_breaks = seq(0, 70, 10),
  value_limits = c(0, 70),
  width        = 0.85,
  title = "Zonnepanelen naar inkomen",
  ylab  = "inkomensgroepen",
  xlab  = "aandeel binnen inkomensgroep (%)")
```

Note `value_breaks`: custom breaks for the value axis go through the
wrapper, not through a second `scale_y_continuous()`, which would
silently replace the wrapper's percentage labels and zero-flush
expansion.

## Automatic vs. manual value-axis limits

Left on its own, the value axis is always flush to the data: its
limits come from `pretty()`'ing whatever range `y` (or `p5`/`p95`,
`lower`/`upper`, ...) actually spans, so the panel starts exactly
where the data does, on both ends. `value_limits` overrides that with
an explicit `c(min, max)` instead -- not a coordinate-system zoom, but
the value axis's own hard limits, drawn flush the same way. Useful to
line several figures up on one shared scale, to leave room for a
reference value the data itself does not reach, or simply to round
the axis to tidier numbers than the data's own extremes happen to
produce. Automatic, on the same `auto` data as above:

```{r col-auto-limits}
cpb_col(auto, x = inkomensgroep, y = share,
  orientation = "horizontal",
  pct_axis    = TRUE,
  width       = 0.6,
  title = "Aandeel huishoudens zonder auto naar inkomen",
  ylab  = "inkomensgroep",
  xlab  = "huishoudens zonder auto")
```

The data itself only spans 13 to 62, so `pretty()` flushes the axis to
0-70 -- already close to the full 0-100% a share is naturally read
against, but not quite it. `value_limits` fixes that instead:

```{r col-manual-limits}
cpb_col(auto, x = inkomensgroep, y = share,
  orientation  = "horizontal",
  pct_axis     = TRUE,
  value_limits = c(0, 100),
  width        = 0.6,
  title = "Aandeel huishoudens zonder auto naar inkomen",
  ylab  = "inkomensgroep",
  xlab  = "huishoudens zonder auto")
```

Nothing about the bars themselves changes; only the axis they are
read against does -- the same bars read as noticeably smaller shares
once the full 0-100% range is behind them. An estimate that fell
*outside* an explicit `value_limits` would be dropped, with a warning,
the same as setting `limits` on any ggplot2 scale -- here every value
already falls well within the wider range, so nothing is cropped.

## A secondary axis

Two series belong in one figure when they are about the same thing
but an order of magnitude apart -- a wealth total in billions, and
the tax raised on it. `sec_y` draws the second series as a line
against a right-hand axis, with `sec_limits` fixing the range that
axis spans. The line is placed by mapping `sec_limits` linearly onto
the primary range, so both axes start together and the line cannot
drift out of step with its own labels.

`sec_ylab`'s caption on the right is placed to line up exactly with
`ylab`'s own caption on the left, but only `save_cpb()` lines it up
exactly; knitr's own plotting device shows an approximate placement
instead (close, not exact, since at print time the plot has not
actually been rendered yet to measure against). So, like every figure
with a `sec_y` below, this one is written to a file with `save_cpb()`
and shown from that file rather than printed directly:

```{r col-secaxis, results = "hide", out.width = "700px"}
nalatenschap <- expand_grid(
  soort = factor(c("waarover aangifte", "waarover geen aangifte"),
                 levels = c("waarover geen aangifte", "waarover aangifte")),
  jaar  = 2007:2022
) |>
  mutate(mld = c(1.6, 2.6, 2.9, 4.4, 4.4, 3.8, 3.9, 4.7,
                 5.2, 5.8, 5.8, 5.6, 5.6, 6.6, 6.4, 6.5,
                 15.8, 15.9, 15.2, 16.0, 15.3, 14.8, 13.9, 13.1,
                 13.9, 15.0, 15.3, 15.5, 16.3, 19.4, 20.5, 22.2))

erfbelasting <- tibble(
  jaar = 2007:2022,
  heffing = c(2.2, 2.1, 1.9, 1.8, 1.6, 1.7, 1.6, 1.5,
              1.6, 1.7, 1.8, 1.8, 1.8, 2.3, 2.4, 2.5)
)

p <- nalatenschap |>
  left_join(erfbelasting, by = "jaar") |>
  cpb_col(x = jaar, y = mld, fill = soort,
    sec_y      = heffing,
    sec_limits = c(0, 3),
    sec_label  = "erfbelasting",
    sec_ylab   = "mld euro",
    title = "Belasting over erfenissen, prijzen 2022",
    ylab  = "mld euro")

path <- tempfile(fileext = ".png")
save_cpb(path, p, page = "full")
```

```{r col-secaxis-show, echo = FALSE, out.width = "700px"}
.show_saved(path)
```

`sec_ylab` mirrors on the right what `ylab` puts above the panel on
the left, and `sec_label` names the line in the legend -- house style
says which axis it belongs to. The combination is refused where it
cannot be read: horizontal bars, grouped categories, and
`position = "fill"` all error rather than draw something misleading.

`cpb_line()` takes the same arguments, for the figure that puts two
rates on the left axis and an index on the right:

```{r line-secondary, results = "hide", out.width = "350px"}
lonen <- tibble(
  jaar  = 2020:2027,
  cao   = c(3.0, 2.2, 3.2, 5.8, 6.7, 5.1, 4.2, 3.8),
  cpi   = c(1.3, 2.7, 10.0, 3.8, 3.3, 3.3, 3.3, 2.8),
  reeel = c(100.5, 100.0, 96.8, 98.5, 99.2, 100.4, 101.6, 102.5)
)

p <- lonen |>
  pivot_longer(c(cao, cpi), names_to = "reeks", values_to = "waarde") |>
  mutate(reeks = recode(reeks,
    cao = "cao-loon bedrijven (linkeras)",
    cpi = "inflatie (cpi, linkeras)")) |>
  cpb_line(x = jaar, y = waarde, colour = reeks,
    sec_y      = reeel,
    sec_limits = c(95, 110),
    sec_label  = "reële cao-loon bedrijven (rechteras)",
    sec_ylab   = "geïndexeerd, 2021 = 100",
    value_limits = c(-5, 10),
    value_breaks = seq(-5, 10, 5),
    forecast_x   = 2025,
    zeroline     = TRUE,
    title = "Cao-loon bedrijven en inflatie (cpi)",
    ylab  = "mutatie in %") +
  scale_x_continuous(breaks = seq(2020, 2026, 2))

path <- tempfile(fileext = ".png")
save_cpb(path, p, page = "half")
```

```{r line-secondary-show, echo = FALSE, out.width = "350px"}
.show_saved(path)
```

The one difference from `cpb_col()`: there the columns key on `fill`
and the secondary line on `colour`, so they form two legend blocks.
Here the primary series already key on `colour`, so the secondary line
joins that same scale, takes the next `index` position and shares a
single legend -- which is why all three labels name their axis.

# Area charts

`cpb_area()` draws the recurring share-of-total-over-time figure.
With `pct_axis = TRUE` the y axis gets Dutch percentage labels:

```{r area}
bronnen <- c("gas", "elektriciteit", "warmte", "overig")
mix <- expand_grid(jaar = 2018:2027,
                   bron = factor(bronnen, levels = bronnen)) |>
  mutate(ruw = runif(n(), 1, 10)) |>
  mutate(aandeel = 100 * ruw / sum(ruw), .by = jaar)

cpb_area(mix, x = jaar, y = aandeel, fill = bron,
  pct_axis = TRUE,
  title    = "Energiemix van huishoudens",
  ylab     = "aandeel") +
  scale_x_continuous(breaks = seq(2018, 2027, 3), minor_breaks = 2018:2027,
                     guide = guide_axis(minor.ticks = TRUE))
```

# Donut charts

`cpb_donut()` draws a single share-of-total breakdown as a ring
instead of a stacked area over time -- one wedge per category, sized
by `y`. There is no value axis to read (nothing to label), so wedge
sizes are read off the legend or the data itself. `ring_width`
controls how thick the ring is, from a thin ring around a large hole
up to `2` (no hole at all -- a full pie).

`cpb_donut()`'s ring keeps a fixed physical size regardless of title
or legend length (see `panel_size`), which knitr's own plotting device
does not apply on its own -- only `save_cpb()` does. So instead of
printing the plot directly, every donut here is written to a file with
`save_cpb()` and shown from that file, exactly as it would look saved
for a report:

```{r donut, results = "hide"}
energie <- tibble(
  bron  = factor(bronnen, levels = bronnen),
  share = c(45, 30, 15, 10)
)

path <- tempfile(fileext = ".png")
save_cpb(path, cpb_donut(energie, fill = bron, y = share,
  index = c(6, 5, 2, 4),
  title = "Energiemix van huishoudens"), page = "half")
```

```{r donut-show, echo = FALSE}
.show_saved(path)
```

With more wedges, printing the value on the wedge itself gets cramped
for the thin slices. `label_style = "leader"` moves every value
outside the ring instead, connected back to its own wedge by a line
that stays perpendicular to the wedge and straight the rest of the
way out -- and automatically spaces out any wedges whose values would
otherwise land on top of each other, like the two thinnest slices
here:

```{r donut-leader, results = "hide"}
niveaus <- c("aardgas", "overig", "windenergie op land", "zonne-energie",
             "kernenergie", "windenergie op zee")
energie2 <- tibble(
  bron  = factor(c("aardgas", "windenergie op land", "overig",
                    "kernenergie", "windenergie op zee", "zonne-energie"),
                 levels = niveaus),
  share = c(40, 35, 13, 3, 2, 7)
)

path <- tempfile(fileext = ".png")
save_cpb(path, cpb_donut(energie2, fill = bron, y = share,
  label_style = "leader",
  index = c(6, 4, 5, 1, 3, 2),
  title = "Energiemix, met lijnlabels"), page = "half")
```

```{r donut-leader-show, echo = FALSE}
.show_saved(path)
```

# Quantile boxplots

`cpb_box()` draws the CPB distributional figure: a box over the
p25--p75 interquartile range with the median at p50, and thin
errorbar whiskers out to p5 and p95. It expects **precomputed
quantile columns** (both layers use `stat = "identity"`), so
aggregate your microdata first:

```{r box-single}
groepen <- c("tot 120% wml", "120% wml - mod.", "1 - 1,5x mod.",
             "1,5 - 2x mod.", "2 - 3x mod.", "boven 3x mod.")
raw <- tibble(
  groep      = factor(rep(groepen, each = 400), levels = groepen),
  koopkracht = rnorm(2400, mean = rep(c(-3, -1.5, 0, 1, 2, 3.5), each = 400), sd = 2)
)
kk <- raw |>
  summarise(
    p5  = quantile(koopkracht, 0.05),
    p25 = quantile(koopkracht, 0.25),
    p50 = quantile(koopkracht, 0.50),
    p75 = quantile(koopkracht, 0.75),
    p95 = quantile(koopkracht, 0.95),
    .by = groep
  )

cpb_box(kk, x = groep,
  p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "horizontal",
  title    = "Koopkracht per inkomensgroep",
  subtitle = "inkomensgroep",
  ylab     = "% koopkrachtmutatie")
```

Boxes without a `fill` mapping are drawn in the CPB primary blue.
The other box constructions -- the legacy and designer styles,
fill-dodged boxes per year, and grouped layouts -- are in
[`vignette("boxplots")`](boxplots.html).

# Dot-and-interval charts

`cpb_dot()` draws the coefficient-plot form used for regression
output: one estimate per row with its interval through it, read
against a black line at zero. It is horizontal by default, so the
rows read top to bottom in the order of the factor levels. `group`
collects the rows under bold headings, exactly as in `cpb_box()`:

```{r dot, fig.width = 5.96, fig.height = 2.98, out.width = "700px"}
termen <- c("Belasting ontmoedigt werken en sparen",
            "Ongelijkheid is oneerlijk", "Succes door hard werken",
            "Vertrouwen in de politiek", "Leeftijdscategorie",
            "Geslacht (vrouw t.o.v. man)", "Heeft kinderen",
            "Vermogenskwintiel")
schatting <- tibble(
  term = factor(termen, levels = termen),
  blok = factor(rep(c("Percepties", "Achtergrond"), each = 4),
                levels = c("Percepties", "Achtergrond")),
  coef = c(3.2, 2.9, -2.0, 2.9, 3.5, 0.4, -1.4, -2.5),
  lo   = c(-0.3, 2.1, -3.0, 1.9, 2.6, -1.3, -3.3, -3.2),
  hi   = c(6.6, 3.6, -1.1, 3.9, 4.3, 2.2, 0.6, -1.8)
)

cpb_dot(schatting, x = term, y = coef, lower = lo, upper = hi,
  group = blok,
  title = "Samenhang voorkeurstarieven, percepties en kenmerken",
  xlab  = "%-punt verandering in voorkeurstarief")
```

The estimates carry no `colour` mapping here, so everything is drawn
in the house pink and no legend appears. Map `colour` to compare two
specifications side by side, and use `cap_width = 0` for plain
intervals without end caps.

# Scatter plots

`cpb_scatter()` draws points in the house style. Without a `colour`
column the points are CPB blue; a *numeric* `colour` column gets the
continuous CPB gradient, a discrete one the discrete palette.

Binning a numeric variable with `cpb_cut()` takes the second route:
the classes get the light-to-dark `"blues"` palette and a legend of
discrete keys, which is easier to read off a dense scatter than a
gradient. `legend_ncol = 2` lays those keys out in two columns rather
than one tall column:

```{r scatter, fig.width = 5.96, fig.height = 2.98, out.width = "700px"}
hh <- tibble(inkomen = round(rlnorm(400, log(2500), 0.35))) |>
  mutate(energierekening = round(90 + 0.04 * inkomen + rnorm(n(), 0, 35)),
         koopkracht      = round(rnorm(n(), (inkomen - 2500) / 1500, 2), 1))

cpb_scatter(hh, x = inkomen, y = energierekening,
  colour = cpb_cut(koopkracht, c(-Inf, -2, -1, 0, 1, 2, Inf),
                   labeller = label_pct_nl()),
  palette   = "blues",
  legend_ncol = 2,
  title = "Energierekening naar inkomen",
  ylab  = "energierekening (euro per maand)",
  xlab  = "besteedbaar inkomen (euro per maand)",
  colourlab = "koopkracht (%)") +
  scale_x_continuous(labels = label_euro_nl())
```

# Histograms

`cpb_hist()` bins a column of observations into house-blue bars with
white outlines; set `binwidth` or `bins`. Map `fill` for grouped
histograms:

```{r hist}
duur <- tibble(maanden = round(rgamma(1200, 8, 0.6)))

cpb_hist(duur, x = maanden, binwidth = 2,
  title = "Verdeling van de duur",
  ylab  = "aantal",
  xlab  = "duur (maanden)")
```

For a rendered gallery of all chart types against the published
reference figures, run `inst/examples/smoke_test_plots.R`.

# Sources

Every data set above is **simulated** and every label is a
placeholder -- nothing here reproduces a published number. What the
examples do follow is the *layout* of real CPB figures, so that the
house style has something concrete to be judged against.

The figures used as those targets come from:

- [Erfbelasting in beeld: feiten, percepties en voorkeuren van
  Nederlanders](https://www.cpb.nl/publicatie/erfbelasting-beeld-feiten-percepties-en-voorkeuren-van-nederlanders)
  -- the layouts behind *Voorkeurstarieven erfbelasting naar
  leeftijd* (markers on the line), *Belasting over erfenissen*
  (secondary axis) and *Samenhang voorkeurstarieven, percepties en
  kenmerken* (dot-and-interval).
- [Inkomenseffecten van hogere energie- en
  brandstofprijzen](https://www.cpb.nl/publicatie/inkomenseffecten-van-hogere-energie-en-brandstofprijzen)
  -- the household-level income and energy figures.

The rendered originals live in `references/plots/` in the repository,
named after the publication they came from.
