---
title: "Chart types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chart types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Figures are drawn on the REAL CPB canvas (page = "half" is 2.98 in,
# "full" is 5.96 in; text sizes in theme_cpb() are absolute points, so
# any other canvas distorts the text-to-figure proportion) and only
# *displayed* smaller via out.width, at a consistent 2:1 scale:
# half-page figures at 350px, full-page figures at 700px.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(42)
```

One section per CPB chart type, each in its *default* house style:
the simulated data it needs and the wrapper call that draws it,
nothing more. All wrappers return a plain `ggplot` object, so
anything here can be extended further with `+`.

Variants and extras live in the other vignettes:
[`vignette("layout")`](layout.html) for facets and grouped category
axes, [`vignette("boxplots")`](boxplots.html) for the box-plot styles
and combinations, [`vignette("annotation")`](annotation.html) for
reference lines and forecast windows,
[`vignette("maps")`](maps.html) for choropleths, and
[`vignette("ggcpb")`](ggcpb.html) for the basic setup and export.

Two house conventions to know up front:

* **Titles and units.** `title` is the bold heading. The value-axis
  unit goes in `ylab` and is rendered as the italic caption above the
  panel (the CPB "subtitle" position), not as a rotated axis title. A
  titled figure always reserves that caption line, so figures with and
  without a unit align.
* **Zero line.** A solid black zero line is drawn automatically on
  the value axis: always for columns and areas (which are anchored at
  zero), and for lines and boxplots whenever the data spans zero.

# Line charts

A single series needs only `x` and `y`; it is drawn in the CPB
primary blue:

```{r line-single}
bbp <- tibble(
  jaar  = 2015:2027,
  index = 100 + cumsum(rnorm(13, mean = 1.5, sd = 1.2))
)

cpb_line(bbp, x = jaar, y = index,
  title = "Bruto binnenlands product",
  ylab  = "index (2015 = 100)") +
  scale_x_continuous(breaks = seq(2015, 2027, 3), minor_breaks = 2015:2027,
                     guide = guide_axis(minor.ticks = TRUE))
```

Map a column to `colour` for multiple series, and pick house colours
by palette position with `index` -- `c(6, 2)` is the recurring
blue/magenta pair. Growth rates span zero, so the black zero line
appears by itself:

```{r line-multi}
groei <- expand_grid(reeks = c("arbeidsproductiviteit", "tfp"),
                     jaar  = 2000:2024) |>
  mutate(waarde = round(rnorm(n(), mean = 1, sd = 1.6), 1))

cpb_line(groei, x = jaar, y = waarde, colour = reeks,
  index = c(6, 2),
  title = "Productiviteitsgroei",
  ylab  = "%") +
  scale_x_continuous(
    breaks       = seq(2000, 2024, 4),
    minor_breaks = 2000:2024,
    guide        = guide_axis(minor.ticks = TRUE)
  )
```

The added `scale_x_continuous()` shows the usual refinements for a
year axis: labelled breaks every few years plus small minor ticks for
the years in between.

## Markers on the line

A dense time series is drawn as a bare line. When the x axis instead
holds a handful of ordered categories -- age brackets, quintiles --
the house style puts a marker on every observation, so the reader can
see where the measurements actually are. `points = TRUE` adds them in
the line colour, and the legend key becomes a line with a marker on
it:

```{r line-points}
leeftijden <- c("20-30", "30-40", "40-50", "50-60", "60-70", "70-80", ">80")
tarieven <- expand_grid(
  erfenis  = factor(c("10.000 euro", "100.000 euro", "1.000.000 euro"),
                    levels = c("10.000 euro", "100.000 euro", "1.000.000 euro")),
  leeftijd = factor(leeftijden, levels = leeftijden)
) |>
  mutate(tarief = c(14, 12, 13, 10, 8, 10, 12,
                    21, 18, 18, 17, 17, 21, 24,
                    29, 27, 28, 27, 29, 37, 41))

cpb_line(tarieven, x = leeftijd, y = tarief, colour = erfenis,
  points       = TRUE,
  index        = c(6, 2, 8),
  value_limits = c(0, 40),
  title = "Voorkeurstarieven erfbelasting naar leeftijd",
  ylab  = "% gemiddeld voorkeurstarief")
```

`points = TRUE` is the only thing separating this from an ordinary
`cpb_line()` call; `point_size` adjusts the marker if the default
reads too heavy against a thicker line.

# Column charts

`cpb_col()` covers stacked, dodged and horizontal bars. Stacked is
the default `position`:

```{r col-stacked}
sectoren <- c("industrie", "diensten", "landbouw", "overheid")
tw <- expand_grid(jaar   = 2023:2027,
                  sector = factor(sectoren, levels = sectoren)) |>
  mutate(waarde = round(runif(n(), 5, 25), 1))

cpb_col(tw, x = jaar, y = waarde, fill = sector,
  index = c(6, 5, 2, 4),
  title = "Toegevoegde waarde per sector",
  ylab  = "mld euro")
```

By default the fill legend is reversed (`reverse_legend = TRUE`), so
it reads in the same top-to-bottom order as the stack. For dodged
bars, switch that off so the legend follows the series order:

```{r col-dodged}
scenario <- expand_grid(regio    = c("Noord", "Oost", "Zuid", "West"),
                        scenario = c("basispad", "beleidsvariant")) |>
  mutate(effect = round(runif(n(), -2, 6), 1))

cpb_col(scenario, x = regio, y = effect, fill = scenario,
  position = "dodge",
  index = c(6, 2),
  reverse_legend = FALSE,
  title = "Effect per regio en scenario",
  ylab  = "% mutatie")
```

## Horizontal bars

`orientation = "horizontal"` flips the chart and moves the gridlines
to the value axis. Here `ylab` labels the *category* axis (still the
caption above the panel) and `xlab` labels the value axis at the
bottom. `pct_axis` formats the value axis with Dutch percentage
labels, `value_limits` fixes its range, and the panel starts exactly
at the zero axis:

```{r col-horizontal}
groepen <- c("tot 120% wml", "120% wml - mod.", "1 - 1,5x mod.",
             "1,5 - 2x mod.", "2 - 3x mod.", "boven 3x mod.")
auto <- tibble(
  # reversed levels put the first-named group at the top after the flip
  inkomensgroep = factor(groepen, levels = rev(groepen)),
  share         = c(62, 35, 26, 19, 14, 13)
)

cpb_col(auto, x = inkomensgroep, y = share,
  orientation  = "horizontal",
  pct_axis     = TRUE,
  value_limits = c(0, 70),
  width        = 0.6,
  title = "Aandeel huishoudens zonder auto naar inkomen",
  ylab  = "inkomensgroep",
  xlab  = "huishoudens zonder auto")
```

A horizontal *dodged* bar needs one extra trick. Under `coord_flip()`
the dodge draws the *last* factor level on top within each group, so
to show (say) 2021 above 2024 you reverse the `jaar` levels, swap the
`index` to keep each year its colour, and let `reverse_legend`
restore the reading order:

```{r col-horizontal-dodged}
pv <- expand_grid(
  jaar          = factor(c(2021, 2024), levels = c(2024, 2021)),
  inkomensgroep = factor(groepen, levels = rev(groepen))
) |>
  mutate(share = c(10, 14, 19, 22, 26, 33,   # 2021
                   19, 25, 31, 38, 47, 59))  # 2024

cpb_col(pv, x = inkomensgroep, y = share, fill = jaar,
  position     = "dodge",
  orientation  = "horizontal",
  index        = c(2, 6),        # level order is (2024, 2021)
  value_breaks = seq(0, 70, 10),
  value_limits = c(0, 70),
  width        = 0.85,
  title = "Zonnepanelen naar inkomen",
  ylab  = "inkomensgroepen",
  xlab  = "aandeel binnen inkomensgroep (%)")
```

Note `value_breaks`: custom breaks for the value axis go through the
wrapper, not through a second `scale_y_continuous()`, which would
silently replace the wrapper's percentage labels and zero-flush
expansion.

## A secondary axis

Two series belong in one figure when they are about the same thing
but an order of magnitude apart -- a wealth total in billions, and
the tax raised on it. `sec_y` draws the second series as a line
against a right-hand axis, with `sec_limits` fixing the range that
axis spans. The line is placed by mapping `sec_limits` linearly onto
the primary range, so both axes start together and the line cannot
drift out of step with its own labels:

```{r col-secaxis, fig.width = 5.96, fig.height = 2.98, out.width = "700px"}
nalatenschap <- expand_grid(
  soort = factor(c("waarover aangifte", "waarover geen aangifte"),
                 levels = c("waarover geen aangifte", "waarover aangifte")),
  jaar  = 2007:2022
) |>
  mutate(mld = c(1.6, 2.6, 2.9, 4.4, 4.4, 3.8, 3.9, 4.7,
                 5.2, 5.8, 5.8, 5.6, 5.6, 6.6, 6.4, 6.5,
                 15.8, 15.9, 15.2, 16.0, 15.3, 14.8, 13.9, 13.1,
                 13.9, 15.0, 15.3, 15.5, 16.3, 19.4, 20.5, 22.2))

erfbelasting <- tibble(
  jaar = 2007:2022,
  heffing = c(2.2, 2.1, 1.9, 1.8, 1.6, 1.7, 1.6, 1.5,
              1.6, 1.7, 1.8, 1.8, 1.8, 2.3, 2.4, 2.5)
)

nalatenschap |>
  left_join(erfbelasting, by = "jaar") |>
  cpb_col(x = jaar, y = mld, fill = soort,
    sec_y      = heffing,
    sec_limits = c(0, 3),
    sec_label  = "erfbelasting (rechteras)",
    sec_ylab   = "mld euro",
    index      = c(8, 6),
    title = "Belasting over erfenissen, prijzen 2022",
    ylab  = "mld euro")
```

`sec_ylab` mirrors on the right what `ylab` puts above the panel on
the left, and `sec_label` names the line in the legend -- house style
says which axis it belongs to. The combination is refused where it
cannot be read: horizontal bars, grouped categories, and
`position = "fill"` all error rather than draw something misleading.

`cpb_line()` takes the same arguments, for the figure that puts two
rates on the left axis and an index on the right:

```{r line-secondary, fig.width = 2.98, fig.height = 3.0}
lonen <- tibble(
  jaar  = 2020:2027,
  cao   = c(3.0, 2.2, 3.2, 5.8, 6.7, 5.1, 4.2, 3.8),
  cpi   = c(1.3, 2.7, 10.0, 3.8, 3.3, 3.3, 3.3, 2.8),
  reeel = c(100.5, 100.0, 96.8, 98.5, 99.2, 100.4, 101.6, 102.5)
)

lonen |>
  pivot_longer(c(cao, cpi), names_to = "reeks", values_to = "waarde") |>
  mutate(reeks = recode(reeks,
    cao = "cao-loon bedrijven (linkeras)",
    cpi = "inflatie (cpi, linkeras)")) |>
  cpb_line(x = jaar, y = waarde, colour = reeks,
    sec_y      = reeel,
    sec_limits = c(95, 110),
    sec_label  = "reële cao-loon bedrijven (rechteras)",
    sec_ylab   = "geïndexeerd, 2021 = 100",
    index        = c(6, 2, 8),
    value_limits = c(-5, 10),
    value_breaks = seq(-5, 10, 5),
    forecast_x   = 2025,
    zeroline     = TRUE,
    title = "Cao-loon bedrijven en inflatie (cpi)",
    ylab  = "mutatie in %") +
  scale_x_continuous(breaks = seq(2020, 2026, 2))
```

The one difference from `cpb_col()`: there the columns key on `fill`
and the secondary line on `colour`, so they form two legend blocks.
Here the primary series already key on `colour`, so the secondary line
joins that same scale, takes the next `index` position and shares a
single legend -- which is why all three labels name their axis.

# Area charts

`cpb_area()` draws the recurring share-of-total-over-time figure.
With `pct_axis = TRUE` the y axis gets Dutch percentage labels:

```{r area}
bronnen <- c("gas", "elektriciteit", "warmte", "overig")
mix <- expand_grid(jaar = 2018:2027,
                   bron = factor(bronnen, levels = bronnen)) |>
  mutate(ruw = runif(n(), 1, 10)) |>
  mutate(aandeel = 100 * ruw / sum(ruw), .by = jaar)

cpb_area(mix, x = jaar, y = aandeel, fill = bron,
  pct_axis = TRUE,
  index    = c(6, 5, 2, 4),
  title    = "Energiemix van huishoudens",
  ylab     = "aandeel") +
  scale_x_continuous(breaks = seq(2018, 2027, 3), minor_breaks = 2018:2027,
                     guide = guide_axis(minor.ticks = TRUE))
```

# Quantile boxplots

`cpb_box()` draws the CPB distributional figure: a box over the
p25--p75 interquartile range with the median at p50, and thin
errorbar whiskers out to p5 and p95. It expects **precomputed
quantile columns** (both layers use `stat = "identity"`), so
aggregate your microdata first:

```{r box-single}
groepen <- c("tot 120% wml", "120% wml - mod.", "1 - 1,5x mod.",
             "1,5 - 2x mod.", "2 - 3x mod.", "boven 3x mod.")
raw <- tibble(
  groep      = factor(rep(groepen, each = 400), levels = groepen),
  koopkracht = rnorm(2400, mean = rep(c(-3, -1.5, 0, 1, 2, 3.5), each = 400), sd = 2)
)
kk <- raw |>
  summarise(
    p5  = quantile(koopkracht, 0.05),
    p25 = quantile(koopkracht, 0.25),
    p50 = quantile(koopkracht, 0.50),
    p75 = quantile(koopkracht, 0.75),
    p95 = quantile(koopkracht, 0.95),
    .by = groep
  )

cpb_box(kk, x = groep,
  p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "horizontal",
  title    = "Koopkracht per inkomensgroep",
  subtitle = "inkomensgroep",
  ylab     = "% koopkrachtmutatie")
```

Boxes without a `fill` mapping are drawn in the CPB primary blue.
The other box constructions -- the legacy and designer styles,
fill-dodged boxes per year, and grouped layouts -- are in
[`vignette("boxplots")`](boxplots.html).

# Dot-and-interval charts

`cpb_dot()` draws the coefficient-plot form used for regression
output: one estimate per row with its interval through it, read
against a black line at zero. It is horizontal by default, so the
rows read top to bottom in the order of the factor levels. `group`
collects the rows under bold headings, exactly as in `cpb_box()`:

```{r dot, fig.width = 5.96, fig.height = 2.98, out.width = "700px"}
termen <- c("Belasting ontmoedigt werken en sparen",
            "Ongelijkheid is oneerlijk", "Succes door hard werken",
            "Vertrouwen in de politiek", "Leeftijdscategorie",
            "Geslacht (vrouw t.o.v. man)", "Heeft kinderen",
            "Vermogenskwintiel")
schatting <- tibble(
  term = factor(termen, levels = termen),
  blok = factor(rep(c("Percepties", "Achtergrond"), each = 4),
                levels = c("Percepties", "Achtergrond")),
  coef = c(3.2, 2.9, -2.0, 2.9, 3.5, 0.4, -1.4, -2.5),
  lo   = c(-0.3, 2.1, -3.0, 1.9, 2.6, -1.3, -3.3, -3.2),
  hi   = c(6.6, 3.6, -1.1, 3.9, 4.3, 2.2, 0.6, -1.8)
)

cpb_dot(schatting, x = term, y = coef, lower = lo, upper = hi,
  group = blok,
  title = "Samenhang voorkeurstarieven, percepties en kenmerken",
  xlab  = "%-punt verandering in voorkeurstarief")
```

The estimates carry no `colour` mapping here, so everything is drawn
in the house pink and no legend appears. Map `colour` to compare two
specifications side by side, and use `cap_width = 0` for plain
intervals without end caps.

# Scatter plots

`cpb_scatter()` draws points in the house style. Without a `colour`
column the points are CPB blue; a *numeric* `colour` column gets the
continuous CPB gradient, a discrete one the discrete palette.

Binning a numeric variable with `cpb_cut()` takes the second route:
the classes get the light-to-dark `"blues"` palette and a legend of
discrete keys, which is easier to read off a dense scatter than a
gradient. `legend_ncol = 2` lays those keys out in two columns rather
than one tall column:

```{r scatter, fig.width = 5.96, fig.height = 2.98, out.width = "700px"}
hh <- tibble(inkomen = round(rlnorm(400, log(2500), 0.35))) |>
  mutate(energierekening = round(90 + 0.04 * inkomen + rnorm(n(), 0, 35)),
         koopkracht      = round(rnorm(n(), (inkomen - 2500) / 1500, 2), 1))

cpb_scatter(hh, x = inkomen, y = energierekening,
  colour = cpb_cut(koopkracht, c(-Inf, -2, -1, 0, 1, 2, Inf),
                   labeller = label_pct_nl()),
  palette   = "blues",
  legend_ncol = 2,
  title = "Energierekening naar inkomen",
  ylab  = "energierekening (euro per maand)",
  xlab  = "besteedbaar inkomen (euro per maand)",
  colourlab = "koopkracht (%)") +
  scale_x_continuous(labels = label_euro_nl())
```

# Histograms

`cpb_hist()` bins a column of observations into house-blue bars with
white outlines; set `binwidth` or `bins`. Map `fill` for grouped
histograms:

```{r hist}
duur <- tibble(maanden = round(rgamma(1200, 8, 0.6)))

cpb_hist(duur, x = maanden, binwidth = 2,
  title = "Verdeling van de duur",
  ylab  = "aantal",
  xlab  = "duur (maanden)")
```

For a rendered gallery of all chart types against the published
reference figures, run `inst/examples/smoke_test_plots.R`.

# Sources

Every data set above is **simulated** and every label is a
placeholder -- nothing here reproduces a published number. What the
examples do follow is the *layout* of real CPB figures, so that the
house style has something concrete to be judged against.

The figures used as those targets come from:

- [Erfbelasting in beeld: feiten, percepties en voorkeuren van
  Nederlanders](https://www.cpb.nl/publicatie/erfbelasting-beeld-feiten-percepties-en-voorkeuren-van-nederlanders)
  -- the layouts behind *Voorkeurstarieven erfbelasting naar
  leeftijd* (markers on the line), *Belasting over erfenissen*
  (secondary axis) and *Samenhang voorkeurstarieven, percepties en
  kenmerken* (dot-and-interval).
- [Inkomenseffecten van hogere energie- en
  brandstofprijzen](https://www.cpb.nl/publicatie/inkomenseffecten-van-hogere-energie-en-brandstofprijzen)
  -- the household-level income and energy figures.

The rendered originals live in `references/plots/` in the repository,
named after the publication they came from.
