---
title: "CPB chart types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CPB chart types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Figures are drawn on the REAL CPB canvas (page = "half" is 2.98 in,
# "full" is 5.96 in; text sizes in theme_cpb() are absolute points, so
# any other canvas distorts the text-to-figure proportion) and only
# *displayed* smaller via out.width, at a consistent 2:1 scale:
# half-page figures at 350px, full-page figures at 700px.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(42)
```

This vignette is a cookbook: one section per CPB chart type, each with
the simulated data it needs and the wrapper call that draws it. All
wrappers return a plain `ggplot` object, so anything here can be
extended further with `+`. See `vignette("ggcpb")` for the
composable core the wrappers build on.

Two house conventions to know up front:

* **Titles and units.** `title` is the bold heading. The value-axis
  unit goes in `ylab` and is rendered as the italic caption above the
  panel (the CPB "subtitle" position), not as a rotated axis title. A
  titled figure always reserves that caption line, so figures with and
  without a unit align.
* **Zero line.** A solid black zero line is drawn automatically on
  the value axis: always for columns and areas (which are anchored at
  zero), and for lines and boxplots whenever the data spans zero.

# Line charts

A single series needs only `x` and `y`; it is drawn in the CPB
primary blue:

```{r line-single}
bbp <- tibble(
  jaar  = 2015:2027,
  index = 100 + cumsum(rnorm(13, mean = 1.5, sd = 1.2))
)

cpb_line(bbp, x = jaar, y = index,
  title = "Bruto binnenlands product",
  ylab  = "index (2015 = 100)") +
  scale_x_continuous(breaks = seq(2015, 2027, 3), minor_breaks = 2015:2027,
                     guide = guide_axis(minor.ticks = TRUE))
```

Map a column to `colour` for multiple series, and pick house colours
by palette position with `index` -- `c(6, 2)` is the recurring
blue/magenta pair. Growth rates span zero, so the black zero line
appears by itself:

```{r line-multi}
groei <- expand_grid(reeks = c("arbeidsproductiviteit", "tfp"),
                     jaar  = 2000:2024) |>
  mutate(waarde = round(rnorm(n(), mean = 1, sd = 1.6), 1))

cpb_line(groei, x = jaar, y = waarde, colour = reeks,
  index = c(6, 2),
  title = "Productiviteitsgroei",
  ylab  = "%") +
  scale_x_continuous(
    breaks       = seq(2000, 2024, 4),
    minor_breaks = 2000:2024,
    guide        = guide_axis(minor.ticks = TRUE)
  )
```

The added `scale_x_continuous()` shows the usual refinements for a
year axis: labelled breaks every few years plus small minor ticks for
the years in between.

# Column charts

`cpb_col()` covers stacked, dodged and horizontal bars. Stacked is
the default `position`:

```{r col-stacked}
sectoren <- c("industrie", "diensten", "landbouw", "overheid")
tw <- expand_grid(jaar   = 2023:2027,
                  sector = factor(sectoren, levels = sectoren)) |>
  mutate(waarde = round(runif(n(), 5, 25), 1))

cpb_col(tw, x = jaar, y = waarde, fill = sector,
  index = c(6, 5, 2, 4),
  title = "Toegevoegde waarde per sector",
  ylab  = "mld euro")
```

By default the fill legend is reversed (`reverse_legend = TRUE`), so
it reads in the same top-to-bottom order as the stack. For dodged
bars, switch that off so the legend follows the series order:

```{r col-dodged}
scenario <- expand_grid(regio    = c("Noord", "Oost", "Zuid", "West"),
                        scenario = c("basispad", "beleidsvariant")) |>
  mutate(effect = round(runif(n(), -2, 6), 1))

cpb_col(scenario, x = regio, y = effect, fill = scenario,
  position = "dodge",
  index = c(6, 2),
  reverse_legend = FALSE,
  title = "Effect per regio en scenario",
  ylab  = "% mutatie")
```

## Horizontal bars

`orientation = "horizontal"` flips the chart and moves the gridlines
to the value axis. Here `ylab` labels the *category* axis (still the
caption above the panel) and `xlab` labels the value axis at the
bottom. `pct_axis` formats the value axis with Dutch percentage
labels, `value_limits` fixes its range, and the panel starts exactly
at the zero axis:

```{r col-horizontal}
groepen <- c("tot 120% wml", "120% wml - mod.", "1 - 1,5x mod.",
             "1,5 - 2x mod.", "2 - 3x mod.", "boven 3x mod.")
auto <- tibble(
  # reversed levels put the first-named group at the top after the flip
  inkomensgroep = factor(groepen, levels = rev(groepen)),
  share         = c(62, 35, 26, 19, 14, 13)
)

cpb_col(auto, x = inkomensgroep, y = share,
  orientation  = "horizontal",
  pct_axis     = TRUE,
  value_limits = c(0, 70),
  width        = 0.6,
  title = "Aandeel huishoudens zonder auto naar inkomen",
  ylab  = "inkomensgroep",
  xlab  = "huishoudens zonder auto")
```

A horizontal *dodged* bar needs one extra trick. Under `coord_flip()`
the dodge draws the *last* factor level on top within each group, so
to show (say) 2021 above 2024 you reverse the `jaar` levels, swap the
`index` to keep each year its colour, and let `reverse_legend`
restore the reading order:

```{r col-horizontal-dodged}
pv <- expand_grid(
  jaar          = factor(c(2021, 2024), levels = c(2024, 2021)),
  inkomensgroep = factor(groepen, levels = rev(groepen))
) |>
  mutate(share = c(10, 14, 19, 22, 26, 33,   # 2021
                   19, 25, 31, 38, 47, 59))  # 2024

cpb_col(pv, x = inkomensgroep, y = share, fill = jaar,
  position     = "dodge",
  orientation  = "horizontal",
  index        = c(2, 6),        # level order is (2024, 2021)
  value_breaks = seq(0, 70, 10),
  value_limits = c(0, 70),
  width        = 0.85,
  title = "Zonnepanelen naar inkomen",
  ylab  = "inkomensgroepen",
  xlab  = "aandeel binnen inkomensgroep (%)")
```

Note `value_breaks`: custom breaks for the value axis go through the
wrapper, not through a second `scale_y_continuous()`, which would
silently replace the wrapper's percentage labels and zero-flush
expansion.

# Area charts

`cpb_area()` draws the recurring share-of-total-over-time figure.
With `pct_axis = TRUE` the y axis gets Dutch percentage labels:

```{r area}
bronnen <- c("gas", "elektriciteit", "warmte", "overig")
mix <- expand_grid(jaar = 2018:2027,
                   bron = factor(bronnen, levels = bronnen)) |>
  mutate(ruw = runif(n(), 1, 10)) |>
  mutate(aandeel = 100 * ruw / sum(ruw), .by = jaar)

cpb_area(mix, x = jaar, y = aandeel, fill = bron,
  pct_axis = TRUE,
  index    = c(6, 5, 2, 4),
  title    = "Energiemix van huishoudens",
  ylab     = "aandeel") +
  scale_x_continuous(breaks = seq(2018, 2027, 3), minor_breaks = 2018:2027,
                     guide = guide_axis(minor.ticks = TRUE))
```

# Quantile boxplots

`cpb_box()` draws the CPB distributional figure: a box over the
p25--p75 interquartile range with the median at p50, and thin
errorbar whiskers out to p5 and p95. It expects **precomputed
quantile columns** (both layers use `stat = "identity"`), so
aggregate your microdata first:

```{r box-single}
groepen5 <- c("laagste 20%", "2e 20%", "midden 20%", "4e 20%", "hoogste 20%")
raw <- tibble(
  groep      = factor(rep(groepen5, each = 400), levels = groepen5),
  koopkracht = rnorm(2000, mean = rep(c(-3, -1.5, 0, 1.5, 3.5), each = 400), sd = 2)
)
kk <- raw |>
  summarise(
    p5  = quantile(koopkracht, 0.05),
    p25 = quantile(koopkracht, 0.25),
    p50 = quantile(koopkracht, 0.50),
    p75 = quantile(koopkracht, 0.75),
    p95 = quantile(koopkracht, 0.95),
    .by = groep
  )

cpb_box(kk, x = groep,
  p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "horizontal",
  title    = "Koopkracht per inkomensgroep",
  subtitle = "inkomensgroep",
  ylab     = "% koopkrachtmutatie")
```

Boxes without a `fill` mapping are drawn in the CPB primary blue.
Map `fill` and pass a `position_dodge()` for grouped boxes -- for
example one pair of years per income group:

```{r box-dodged, fig.width = 5.96, fig.height = 2.98, out.width = "700px"}
kk2 <- expand_grid(kk, jaar = factor(c(2026, 2027))) |>
  mutate(across(p5:p95, \(q) q + (jaar == "2027") * 0.8))

cpb_box(kk2, x = groep,
  p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  fill     = jaar,
  position = position_dodge(width = 0.6),
  index    = c(6, 2),
  title    = "Koopkracht per jaar, 2026 en 2027",
  ylab     = "% koopkrachtmutatie") +
  scale_y_continuous(labels = label_number_nl())
```

# Composing: a line over stacked columns

Because every wrapper returns a real `ggplot`, a decomposition chart
-- stacked contribution columns with the total drawn as a line -- is
just `cpb_col()` plus a `geom_line()` layer and a colour scale for
the extra series:

```{r overlay}
componenten <- c("kapitaal/uren", "arbeidssamenstelling", "tfp")
dec <- expand_grid(jaar = 2000:2024,
                   # reversed levels stack the first component nearest zero
                   component = factor(componenten, levels = rev(componenten))) |>
  mutate(bijdrage = round(rnorm(n(), mean = 0.4, sd = 0.9), 1))
totaal <- dec |>
  summarise(bijdrage = sum(bijdrage), .by = jaar)

cpb_col(dec, x = jaar, y = bijdrage, fill = component,
  position = "stack",
  index    = c(2, 5, 6),
  width    = 0.75,
  title    = "Opbouw productiviteitsgroei",
  ylab     = "%") +
  geom_line(
    data    = totaal,
    mapping = aes(x = jaar, y = bijdrage, colour = "arbeidsproductiviteit"),
    linewidth = 0.55, inherit.aes = FALSE
  ) +
  scale_colour_cpb_manual(index = 1) +
  labs(colour = NULL) +
  guides(fill   = guide_legend(reverse = TRUE, order = 1),
         colour = guide_legend(order = 2)) +
  theme(legend.box = "horizontal", legend.box.just = "top")
```

# Everything is a ggplot object

The wrappers do not draw anything -- they *return* a `ggplot` object
with the theme and scales already applied. That means the full
`ggplot2` grammar stays available through `+`: extra geoms,
annotations, scale tweaks and further `theme()` overrides all layer
on top of what the wrapper built.

A dashed reference line with an italic annotation, on top of the
horizontal bar chart from before:

```{r layer-refline}
cpb_col(auto, x = inkomensgroep, y = share,
  orientation  = "horizontal",
  pct_axis     = TRUE,
  value_limits = c(0, 70),
  width        = 0.6,
  title = "Aandeel huishoudens zonder auto naar inkomen",
  ylab  = "inkomensgroep",
  xlab  = "huishoudens zonder auto") +
  geom_hline(yintercept = 24, linetype = "dashed",
             colour = "#666666", linewidth = 0.4) +
  annotate("text", x = 6.45, y = 24, label = "landelijk gemiddelde",
           hjust = -0.05, size = 2.0, colour = "#666666",
           family = cpb_font_family(), fontface = "italic")
```

Note that under `coord_flip()` the value axis is still the `y`
aesthetic, so a reference line on the value axis is a `geom_hline()`.

Or mark a forecast window on a line chart with a shaded region, as
the published scenario figures do:

```{r layer-raming}
cpb_line(bbp, x = jaar, y = index,
  title = "Bruto binnenlands product",
  ylab  = "index (2015 = 100)") +
  annotate("rect", xmin = 2024.5, xmax = Inf, ymin = -Inf, ymax = Inf,
           fill = "white", alpha = 0.4) +
  annotate("text", x = 2026, y = 104, label = "raming",
           size = 2.2, colour = "#666666",
           family = cpb_font_family(), fontface = "italic") +
  geom_vline(xintercept = 2024.5, linetype = "dashed", linewidth = 0.3) +
  scale_x_continuous(breaks = seq(2015, 2027, 3), minor_breaks = 2015:2027,
                     guide = guide_axis(minor.ticks = TRUE))
```

Later `+ theme(...)` calls override individual elements of
`theme_cpb()` the same way -- see the per-figure margin and legend
tweaks in `inst/examples/smoke_test_plots.R`. The one thing *not* to
add is a second value-axis scale: use the wrapper's `value_breaks`,
`value_limits` and `pct_axis` arguments instead, as noted above.

# Export

`save_cpb()` writes the figure at the strict CPB page widths --
`page = "half"` (2.98 in) or `page = "full"` (5.96 in) -- through the
`ragg` device, so the bundled Rijksoverheid font renders correctly:

```{r save, eval = FALSE}
p <- cpb_line(bbp, x = jaar, y = index,
  title = "Bruto binnenlands product",
  ylab  = "index (2015 = 100)")

save_cpb("bbp.png", p, page = "half")
save_cpb("bbp_breed.png", p, page = "full", height = 3.2)
```

For a rendered gallery of all chart types against the published
reference figures, run `inst/examples/smoke_test_plots.R`.
