---
title: "Colours"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colours}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(42)
```

How series get their colours, and how to override them. Everything
here applies equally to `cpb_line()` and `cpb_col()`; the only
difference is the argument name, which follows the aesthetic being
coloured:

* `colour_index` on the wrappers that map `colour` -- `cpb_line()`,
  `cpb_scatter()`, `cpb_dot()`
* `fill_index` on the wrappers that map `fill` -- `cpb_col()`,
  `cpb_area()`, `cpb_box()`, `cpb_hist()`, `cpb_map()`

`color_index` is accepted wherever `colour_index` is.

```{r data}
groei <- expand_grid(
  reeks = c("arbeidsproductiviteit", "tfp", "arbeidsvolume"),
  jaar  = 2015:2024
) |>
  mutate(waarde = round(rnorm(n(), mean = 1, sd = 1.4), 1))

sectoren <- c("industrie", "diensten", "landbouw", "overheid")
tw <- expand_grid(jaar   = 2023:2027,
                  sector = factor(sectoren, levels = sectoren)) |>
  mutate(waarde = round(runif(n(), 5, 25), 1))
```

# The default

Map a column to `colour` or `fill` and the series are coloured for
you, in the order the published figures use: **blue, magenta, taupe**,
then the remaining swatches. No argument needed.

```{r default-line}
cpb_line(groei, x = jaar, y = waarde, colour = reeks,
  title = "Default series colours",
  ylab  = "%")
```

The same order drives fills, so a line chart and a column chart of the
same series agree:

```{r default-col}
cpb_col(tw, x = jaar, y = waarde, fill = sector,
  title = "Default fill colours",
  ylab  = "mld euro")
```

A single series is not part of that cycle: it is drawn in the CPB
primary blue, which is why one-line and one-column figures look the
same as the first series of a multi-series one.

```{r single}
cpb_line(filter(groei, reeks == "tfp"), x = jaar, y = waarde,
  title = "A single series is always primary blue",
  ylab  = "%")
```

# Choosing swatches by position

The house palette is nine swatches. `cpb_cols()` pulls them out by
position, which is how CPB source scripts refer to them:

```{r swatches}
cpb_cols()
```

Pass those positions to `colour_index` / `fill_index` to depart from
the default order -- `c(6, 2)` is the recurring blue/magenta pair,
`8` the taupe used for a third, quieter series:

```{r positions-line}
cpb_line(groei, x = jaar, y = waarde, colour = reeks,
  colour_index = c(7, 5, 4),
  title = "Chosen by position: c(7, 5, 4)",
  ylab  = "%")
```

```{r positions-col}
cpb_col(tw, x = jaar, y = waarde, fill = sector,
  fill_index = c(6, 5, 2, 4),
  title = "Chosen by position: c(6, 5, 2, 4)",
  ylab  = "mld euro")
```

Position numbers stay stable regardless of the default order, so
`cpb_cols(6)` is the primary blue whatever else changes.

## When the order matters

Two cases where you will reach for explicit positions rather than the
default.

**Keeping a series on its colour.** When levels are reordered -- to
control stacking, or because `coord_flip()` draws the last level first
-- the default cycle follows the *level* order, so a series can change
colour between two charts of the same data. Naming positions pins each
series to its colour.

**Matching a figure you are reproducing.** A chart rebuilt from a
publication needs that publication's colours, not the current default.

# Sequential shades

For ordered classes -- income bands, intensity classes -- the discrete
palette is the wrong tool: its colours are meant to be *distinct*, not
ranked. Pass `"continuous"` to switch to the sequential ramp, which
runs light to dark:

```{r continuous-col}
klassen <- c("0-5%", "5-10%", "10-20%", "20-30%", "meer dan 30%")
intensiteit <- expand_grid(jaar   = 2024:2027,
                           klasse = factor(klassen, levels = klassen)) |>
  mutate(n = round(runif(n(), 40, 220)))

cpb_col(intensiteit, x = jaar, y = n, fill = klasse,
  fill_index = "continuous",
  title = "Ordered classes take the ramp",
  ylab  = "aantal personen, x 1000")
```

`"discrete"` names the default explicitly, which is worth doing when a
chart sits next to a `"continuous"` one and you want the contrast to
be visible in the code:

```{r discrete-keyword}
cpb_col(tw, x = jaar, y = waarde, fill = sector,
  fill_index = "discrete",
  title = "The default, named explicitly",
  ylab  = "mld euro")
```

The keywords `"qualitative"`, `"sequential"`, `"discr"` and `"blues"`
name the underlying palettes directly, for the two ramps that have no
short name of their own.

# The palette argument

`palette` selects the same palettes and predates the keywords. Both
still work, so pick one per call: a keyword that contradicts an
explicit `palette` is an error rather than a silent winner.

```{r conflict, error = TRUE}
cpb_col(tw, x = jaar, y = waarde, fill = sector,
  fill_index = "continuous",
  palette    = "qualitative")
```

`palette` is also the way to reach a palette while still picking
positions out of it, since `fill_index` can carry positions *or* a
palette name, not both at once:

```{r palette-plus-positions}
cpb_col(intensiteit, x = jaar, y = n, fill = klasse,
  palette    = "blues",
  fill_index = c(2, 3, 4, 5, 6),
  title = "Positions out of a named palette",
  ylab  = "aantal personen, x 1000")
```

# Colours outside the palette

For a one-off colour that is not a house swatch, the wrappers return
plain `ggplot` objects, so an ordinary scale still applies:

```{r manual-scale}
cpb_line(groei, x = jaar, y = waarde, colour = reeks,
  title = "A manual scale still works",
  ylab  = "%") +
  scale_colour_manual(values = c("#005faf", "#96827d", "#64504b"))
```

For a single series, `line_colour` (and `fill_colour` on the column
wrappers) sets the one colour directly, without involving a scale at
all:

```{r single-colour}
cpb_line(filter(groei, reeks == "tfp"), x = jaar, y = waarde,
  line_colour = cpb_cols(3),
  title = "A single series in a chosen swatch",
  ylab  = "%")
```

# Summary

| Goal | Argument |
| --- | --- |
| House colours, published order | nothing -- it is the default |
| Specific swatches | `colour_index` / `fill_index = c(6, 2)` |
| Ordered classes | `fill_index = "continuous"` |
| Name the default explicitly | `fill_index = "discrete"` |
| Positions out of another palette | `palette =` plus positions |
| One series, one colour | `line_colour` / `fill_colour` |
| Anything else | a plain `scale_*_manual()` |
