---
title: "Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Figures are drawn on the REAL CPB canvas (page = "half" is 2.98 in,
# "full" is 5.96 in) and displayed at a consistent 2:1 scale: half-page
# figures at 350px, full-page figures at 700px.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

ggcpb encodes the CPB house plotting style on a ggplot2 backend. It
has two layers, and this vignette shows the basic idea of both:

* **Wrappers** (`cpb_line()`, `cpb_col()`, `cpb_box()`, ...): one call
  gives a complete house-style figure -- theme, colours, zero line,
  flush legend, Dutch number formats. Every wrapper *returns* a plain
  `ggplot` object, so the full ggplot2 grammar stays available through
  `+`.
* **The composable core** (`theme_cpb()`, the `scale_*_cpb_*()`
  scales, palettes, formatters, fonts): the pieces the wrappers are
  built from, for any figure without a wrapper.

The other vignettes go deeper: `vignette("chart-types")` for the
default chart per type, `vignette("layout")` for facets and grouped
axes, `vignette("annotation")` for reference lines and forecast
windows, `vignette("boxplots")` for the box-plot styles, and
`vignette("maps")` for choropleths of the Netherlands.

```{r setup, message = FALSE}
library(ggcpb)
library(ggplot2)
library(dplyr)
set.seed(7)
```

# One wrapper call

The standard CPB income groups, with a purchasing-power change per
group and year:

```{r data}
groepen <- c("tot 120% wml", "120% wml - mod.", "1 - 1,5x mod.",
             "1,5 - 2x mod.", "2 - 3x mod.", "boven 3x mod.")

kk <- tibble(
  # reversed levels put the first-named group at the top after the flip
  groep      = factor(groepen, levels = rev(groepen)),
  koopkracht = c(-1.2, -0.4, 0.3, 0.8, 1.4, 2.1)
)

p <- cpb_col(kk, x = groep, y = koopkracht,
  orientation = "horizontal",
  title = "Koopkracht naar inkomensgroep",
  ylab  = "inkomensgroep",
  xlab  = "% koopkrachtmutatie")
p
```

One call, one finished figure: bold title, italic caption line, black
hairline gridlines on the value axis, ticks on the category axis, the
black zero line, and the CPB blue -- all house defaults, no styling
arguments needed.

# It is still a ggplot

Because the wrapper returned a real `ggplot` object, refining it is
ordinary ggplot2: `+` a layer, a scale, an annotation. Here a dashed
reference line in the house magenta (`cpb_cols(2)`), labelled in the
bundled house font:

```{r layer}
p +
  geom_hline(yintercept = 0.9, linetype = "dashed",
             colour = cpb_cols(2), linewidth = 0.4) +
  annotate("text", x = 6.4, y = 0.9, label = "raming",
           hjust = -0.15, size = 2.0, colour = cpb_cols(2),
           family = cpb_font_family(), fontface = "italic")
```

(Under `coord_flip()` the value axis is still the `y` aesthetic, so
the reference line is a `geom_hline()`. More annotation patterns in
`vignette("annotation")`.)

# Without a wrapper: the theme and scales directly

Not every figure has a wrapper. For anything else you build from raw
`ggplot2` and apply the same core pieces the wrappers use:
`theme_cpb()`, a CPB colour scale (`scale_fill_cpb_c()` for a
continuous gradient), and the Dutch-locale formatters. Two house
conventions to carry over yourself: the value-axis unit goes in
`subtitle` (never a rotated y-axis title), and the horizontal axis
title sits at the bottom right.

```{r heatmap, fig.width = 5.96, fig.height = 2.4, out.width = "700px"}
raster <- expand.grid(groep = factor(groepen, levels = rev(groepen)),
                      jaar  = 2024:2027) |>
  mutate(mediaan = round(rnorm(n(), 0.5, 1), 1))

ggplot(raster, aes(jaar, groep, fill = mediaan)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  labs(title = "Mediane koopkracht per groep en jaar",
       subtitle = "inkomensgroep",
       x = NULL, y = NULL, fill = "mediaan (%)") +
  scale_fill_cpb_c() +
  theme_cpb(grid = "none", ticks = FALSE)
```

`theme_cpb()` takes the same layout arguments as the wrappers
(`?theme_cpb`), and `cpb_tokens()` exposes the raw design tokens
(palettes, background, grid and NA colours) for anything the scales
do not cover.

# Export

`save_cpb()` writes the figure at the strict CPB page widths --
`page = "half"` (2.98 in) or `page = "full"` (5.96 in) -- through the
`ragg` device, so the bundled Rijksoverheid font (registered
automatically on load; see `cpb_register_fonts()`) renders correctly:

```{r save, eval = FALSE}
save_cpb("koopkracht.png", p, page = "half")
save_cpb("koopkracht_breed.png", p, page = "full", height = 3.2)
```

The half/full widths are the only ones `save_cpb()` accepts: a stray
`width = 8` fails loudly instead of silently producing an off-spec
figure. Text sizes in `theme_cpb()` are absolute points, so the
canvas size is part of the design -- draw at 2.98/5.96 in and scale
the *display*, never the figure. Height defaults to the report
height; pass `height` for taller figures (grouped boxes, facets) or
`preset = "presentation"`.
