---
title: "ggcpb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggcpb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # the real CPB half-page canvas: theme_cpb() text sizes are absolute
  # points, so rendering on a larger canvas shrinks the text relative
  # to the figure; out.width upscales the display instead
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "447px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
library(ggplot2)
library(dplyr)
```

This vignette walks through the composable core of `ggcpb` -- the
theme, the discrete and continuous colour scales, manual palette
selection, the Dutch-locale formatters, and `save_cpb()` -- the
building blocks for styling figures you construct from raw `ggplot2`.
For the high-level chart wrappers (`cpb_line()`, `cpb_col()`,
`cpb_area()`, `cpb_box()`), which apply all of this in one call, see
`vignette("chart-types")`.

# The theme ----

`theme_cpb()` applies the CPB house style on top of
`ggplot2::theme_minimal()`. The default `style = "cpb_default"` is the
look of published CPB figures: hairline black gridlines at labelled
breaks only, black tick marks (with an axis line) on the category
axis, absolute text sizes (9 pt bold title, 7 pt axis text), a
flush-left bottom legend, and the CPB-blue plot background.

```{r theme-basic}
pv_counts <- tibble(
  jaar = 2019:2023,
  aantal = c(120, 135, 128, 150, 162)
)

ggplot(pv_counts, aes(jaar, aantal)) +
  geom_col(fill = cpb_cols(6)) +
  labs(title = "Aantal PV-meldingen per jaar", x = NULL, y = "aantal") +
  theme_cpb()
```

For a chart built with `coord_flip()`, pass `orientation = "horizontal"`
so the gridlines (and ticks) move to the right axes:

```{r theme-horizontal}
ggplot(pv_counts, aes(factor(jaar), aantal)) +
  geom_col(fill = cpb_cols(6)) +
  coord_flip() +
  labs(title = "Aantal PV-meldingen per jaar", x = NULL, y = "aantal") +
  theme_cpb(orientation = "horizontal")
```

`style = "ggplot"` switches to the lighter look of the hand-rolled
CPB ggplot2 scripts: CPB-grey gridlines including minors, no ticks,
6 pt axis text, legend on the right:

```{r theme-ggplot-style}
ggplot(pv_counts, aes(jaar, aantal)) +
  geom_col(fill = cpb_cols(6)) +
  labs(title = "Aantal PV-meldingen per jaar", x = NULL, y = "aantal") +
  theme_cpb(style = "ggplot")
```

Both presets are only defaults. Each element is an individual
argument that overrides the preset: `minor` (minor gridlines),
`ticks` (category-axis tick marks), `grid_colour`/`grid_linewidth`,
`axis_text_size`, `legend`/`flush_legend` (position and the
flush-left bottom block), and `legend_key_size`. `grid` selects which
axes get gridlines (`"value"`, `"both"`, `"none"`, `"x"`, `"y"`), and
`background = FALSE` (or the `theme_cpb_min()` shorthand) drops the
CPB background fill -- useful for small multiples.

# Discrete scales ----

`scale_fill_cpb_d()` / `scale_colour_cpb_d()` draw from one of the
three CPB palettes (`"qualitative"`, `"discr"`, `"sequential"`) and
route `NA` values to the CPB NA colour automatically.

```{r scales-discrete}
pv_by_group <- tibble(
  jaar = rep(2021:2023, each = 2),
  groep = rep(c("huishoudens", "bedrijven"), 3),
  aantal = c(60, 40, 68, 52, 74, 58)
)

ggplot(pv_by_group, aes(jaar, aantal, fill = groep)) +
  geom_col(position = "stack") +
  labs(title = "PV-meldingen naar groep", x = NULL, y = "aantal", fill = NULL) +
  scale_fill_cpb_d() +
  theme_cpb()
```

# Continuous scales ----

`scale_fill_cpb_c()` / `scale_colour_cpb_c()` build a full gradient
across the CPB sequential palette, from its lightest to its darkest
entry.

```{r scales-continuous}
ggplot(mtcars, aes(wt, mpg, colour = hp)) +
  geom_point(size = 2) +
  labs(title = "Gewicht versus verbruik", x = "gewicht", y = "mpg", colour = "pk") +
  scale_colour_cpb_c() +
  theme_cpb(legend = "right")
```

# Manual palette selection ----

`scale_fill_cpb_manual()` / `scale_colour_cpb_manual()` (and the
`cpb_cols()` accessor) select and order specific palette positions,
for the common case of a small, deliberately ordered subset -- e.g.
`index = c(6, 2)` for the recurring CPB blue/magenta pair.

```{r scales-manual}
raming_vergelijking <- tibble(
  scenario = factor(c("basispad", "hoog scenario"), levels = c("basispad", "hoog scenario")),
  effect = c(-1.2, 2.4)
)

ggplot(raming_vergelijking, aes(scenario, effect, fill = scenario)) +
  geom_col() +
  geom_hline(yintercept = 0, colour = "black", linewidth = 0.25) +
  labs(title = "Effect op koopkracht", x = NULL, y = "%-punt", fill = NULL) +
  scale_fill_cpb_manual(index = c(6, 2)) +
  theme_cpb(legend = "none")
```

(The wrappers add that black zero line for you; when composing from
raw `ggplot2` it is a one-line `geom_hline()`.)

# Dutch-locale formatters ----

`label_euro_nl()`, `label_pct_nl()` and `label_number_nl()` wrap
`scales::label_*()` with the Dutch thousands separator (`.`) and
decimal mark (`,`).

```{r formatters}
kosten_vergelijking <- tibble(
  maatregel = factor(c("optie A", "optie B"), levels = c("optie A", "optie B")),
  kosten = c(1250000, 2100000)
)

ggplot(kosten_vergelijking, aes(maatregel, kosten, fill = maatregel)) +
  geom_col() +
  labs(title = "Geraamde kosten per maatregel", x = NULL, y = NULL, fill = NULL) +
  scale_y_continuous(labels = label_euro_nl()) +
  scale_fill_cpb_manual(index = c(6, 2)) +
  theme_cpb(legend = "none")
```

# Raw values and tokens ----

`cpb_tokens()` exposes the raw design tokens (palettes, background,
grid and NA colours) and `cpb_pal()`/`cpb_cols()` generate palette
subsets, for anything the scales above do not cover:

```{r tokens}
cpb_cols(c(6, 2))
cpb_tokens()$bg
```

# Fonts ----

The bundled Rijksoverheid Sans Text family registers automatically
when the package loads; `cpb_font_family()` reports the family name
`theme_cpb()` resolves to (`"RijksoverheidSansText"`, or `""` when
falling back to the ggplot2 default), and `cpb_register_fonts()`
re-runs registration in a fresh worker process.

# Saving figures ----

`save_cpb()` enforces the CPB page widths (half page: 2.98 in, full
page: 5.96 in) and renders with `ragg::agg_png()` by default, so the
bundled CPB font renders correctly.

```{r save, eval = FALSE}
p <- ggplot(pv_counts, aes(jaar, aantal)) +
  geom_col(fill = cpb_cols(6)) +
  labs(title = "Aantal PV-meldingen per jaar", x = NULL, y = "aantal") +
  theme_cpb()

save_cpb("pv_counts.png", p, page = "half")
save_cpb("pv_counts_presentatie.png", p, page = "half", preset = "presentation")
```
