---
title: "Import from CSV"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Import from CSV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Figures are drawn on the REAL CPB canvas (page = "half" is 2.98 in,
# "full" is 5.96 in; text sizes in theme_cpb() are absolute points, so
# any other canvas distorts the text-to-figure proportion) and only
# *displayed* smaller via out.width, at a consistent 2:1 scale:
# half-page figures at 350px, full-page figures at 700px.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
```

Every other vignette builds figures by calling a `cpb_*()` wrapper
directly from R. `import_csv()` is a CSV-only alternative to that: two
plain CSV files -- one with the data, one naming which figures to
build from it -- and no R code at all beyond the one `import_csv()`
call. It exists for the case where the data and the desired figures
live in a spreadsheet workflow rather than an R script -- a colleague
who edits `params.csv` in Excel gets a new or restyled figure without
ever opening R.

# The data

`data_csv` is the plain series data, in long format: one row per
x/series combination, exactly what any `cpb_*()` wrapper's own `data`
argument expects. Here is the example CSV this vignette uses, shipped
with the package so you can open it yourself:

```{r data-path}
data_csv <- system.file("extdata", "import_example_data.csv", package = "ggcpb")
data_csv
```

Its raw contents -- this is exactly what a colleague would see opening
it in a text editor or Excel:

```{r data-raw, echo = FALSE, comment = ""}
cat(readLines(data_csv), sep = "\n")
```

`import_csv()` reads it with `read.csv(check.names = FALSE)`, giving
an ordinary data.frame:

```{r data-frame}
read.csv(data_csv)
```

Four columns: `jaar` and `groep` are the axes/grouping columns,
`koopkracht` is the main series (percentage purchasing-power change),
`werkloosheid` a second series (the unemployment rate) that repeats
across `groep` within a year -- it is one figure's worth of data, not
per group, and a repeated value in a long-format column like this is
completely normal.

# The parameters

`params_csv` names, for each figure, which `cpb_*()` wrapper to call
(`plot_type`) and its arguments. It can be laid out either way --
detected automatically -- and this example uses the **horizontal**
layout: one column per parameter, one row per figure (a spreadsheet's
natural shape). Column-referencing arguments (`x`, `y`, `fill`,
`sec_y`, ...) take a column name from `data_csv`; everything else
(`title`, `ylab`, `position`, ...) is a literal value:

```{r params-path}
params_csv <- system.file("extdata", "import_example_params.csv", package = "ggcpb")
```

```{r params-raw, echo = FALSE, comment = ""}
cat(readLines(params_csv), sep = "\n")
```

Two figures:

* **`koopkracht`** -- `cpb_col()`, `koopkracht` by `jaar`, dodged bars
  coloured by `groep` (`position = "dodge"`; the default `"stack"`
  would sum the three groups' percentages, which does not mean
  anything), plus `werkloosheid` overlaid as a secondary-axis line
  (`sec_y`/`sec_type`/`sec_label`/`sec_ylab`).
* **`werkloosheid`** -- a plain `cpb_line()` of the same series on its
  own.

For the full list of a given `plot_type`'s own parameters (column
reference or literal, and its default), see
`system.file("extdata", "parameter_settings.csv", package = "ggcpb")`
or `?import_csv`.

# Building the figures

One call builds every figure `params_csv` defines, keyed by each
figure's `id`:

```{r build}
# import_csv() (over)writes a run log next to params_csv -- fine for a
# params.csv living in a project folder, but this one lives inside the
# installed package, so the call below points at a throwaway copy
# instead of writing into that installation.
tmp_params <- tempfile(fileext = ".csv")
file.copy(params_csv, tmp_params, overwrite = TRUE)

figs <- import_csv(data_csv, tmp_params)
names(figs)
```

```{r fig-koopkracht, fig.width = 5.96, out.width = "700px"}
figs$koopkracht
```

```{r fig-werkloosheid}
figs$werkloosheid
```

Every call also writes a plain-text run log next to `params_csv`,
listing each figure's outcome (built, skipped and why, or built with
any parameter names it ignored) -- the log is there so a script run
unattended (e.g. from a scheduled task, where the console may not be
visible) still leaves a record:

```{r log, echo = FALSE, comment = ""}
log_path <- paste0(tools::file_path_sans_ext(tmp_params), "_log.txt")
cat(readLines(log_path), sep = "\n")
```

# The vertical layout

The other `params_csv` layout -- **vertical**, meta-tab style -- suits
a single figure with many parameters better: column 1 holds parameter
names, one row each, and column 2 holds that one figure's values. Same
idea, transposed:

```{r vertical}
params_vertical <- tempfile(fileext = ".csv")
writeLines(c(
  "plot_type,line",
  "title,Werkloosheid",
  "x,jaar",
  "y,werkloosheid",
  "ylab,%"
), params_vertical)
cat(readLines(params_vertical), sep = "\n")
```

`params_csv` defining a single figure, `import_csv()` returns that
`ggplot` object directly rather than a named list:

```{r vertical-build}
import_csv(data_csv, params_vertical)
```

# Handling mistakes

A blank cell, or a parameter name that is not a real argument of the
chosen `plot_type`, is simply skipped -- the wrapper's own default
applies, so a typo in `params_csv` never breaks the import (a
`warning()` is still raised, and recorded in the run log). A figure
that fails outright (e.g. a column name that does not exist in
`data_csv`) is skipped the same way, with the actual error recorded --
one broken row in `params_csv` never takes the rest of the batch down
with it. See `?import_csv` for the details.
