---
title: "Building figures from CSV files"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Building figures from CSV files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Figures are drawn on the REAL CPB canvas (page = "half" is 2.98 in,
# "full" is 5.96 in; text sizes in theme_cpb() are absolute points, so
# any other canvas distorts the text-to-figure proportion) and only
# *displayed* smaller via out.width, at a consistent 2:1 scale:
# half-page figures at 350px, full-page figures at 700px.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
```

```{r show-saved-helper, include = FALSE}
# Displays a file save_cpb() already wrote to a temporary path, at a
# stable location knitr manages itself -- so the rendered copy of this
# page (e.g. on GitHub) keeps a working image link rather than one
# pointing at a temp file that no longer exists. Purely a display
# detail for this page, not part of the save_cpb() example above it.
.show_saved <- function(path) {
  stable <- knitr::fig_path(".png")
  dir.create(dirname(stable), recursive = TRUE, showWarnings = FALSE)
  file.copy(path, stable, overwrite = TRUE)
  knitr::include_graphics(stable)
}
```

# Introduction

Most figures in this package are built by calling one of its
functions directly in R code. This vignette describes an alternative
based on two CSV files, a data file and a parameter file.

The data file contains the data used in the figures. The parameter
file specifies which figures to create and how they should be
configured. The `import_csv()` function reads both files and creates
the figures.

This approach is useful when figure specifications need to be kept
separate from R code, for example when the data and figure settings
are maintained outside an R script.

# The data file

The data file can be any CSV file containing the variables needed by
the figures. For this example, the package includes
`import_example_data.csv`.

```{r data-path}
data_csv <- system.file(
  "extdata",
  "import_example_data.csv",
  package = "ggcpb"
)
```

The file can be read in the usual way:

```{r data-frame}
read.csv(data_csv)
```

The important point for the import method is that the parameter file
refers to variables by their column names. For example, if a
parameter contains `jaar`, `import_csv()` looks for a column named
`jaar` in the data file.

# The parameter file

The parameter file determines which figures are created. Each figure
has a `plot_type` and an `id`, together with the settings required
for that plot type.

The example parameter file is `import_example_params.csv`.

```{r params-path}
params_csv <- system.file(
  "extdata",
  "import_example_params.csv",
  package = "ggcpb"
)
```

This example uses the horizontal layout:

```{r params-raw, echo = FALSE, comment = ""}
cat(readLines(params_csv), sep = "\n")
```

The same content, split on the comma and shown as a table:

```{r params-table, echo = FALSE}
knitr::kable(read.csv(params_csv, stringsAsFactors = FALSE))
```

Each row describes one figure. The column names are parameter names.

Some parameters refer to columns in the data file. For example:

* `x`, `y`, `fill` and `sec_y` are column references.
* `title`, `ylab` and `position` are literal values.

The distinction is important. `y,koopkracht` means that the `y`
variable should be taken from the `koopkracht` column, while
`ylab,% mutatie` supplies the text `% mutatie` directly.

# Creating the figures

Once the two files are available, pass them to `import_csv()`:

```{r build, results = "hide"}
tmp_params <- tempfile(fileext = ".csv")
file.copy(params_csv, tmp_params, overwrite = TRUE)

figs <- import_csv(data_csv, tmp_params)
```

When the parameter file contains multiple figures, `import_csv()`
returns a named list. The names come from the `id` parameter.

The `koopkracht` figure has a second value axis (`sec_y`), and its
right side caption is only placed exactly by `save_cpb()`. Printed
directly, an approximate placement is shown instead, close but not
exact, since at that point the figure has not actually been saved yet
to measure against. So, rather than printing it directly, it is
written out through `save_cpb()` first and shown from that file:

```{r fig-koopkracht, results = "hide"}
path <- tempfile(fileext = ".png")
save_cpb(path, figs$koopkracht, page = "full")
```

```{r fig-koopkracht-show, echo = FALSE, out.width = "700px"}
.show_saved(path)
```

`werkloosheid` has no second axis, so it can be shown directly:

```{r fig-werkloosheid}
figs$werkloosheid
```

In normal use, the data and parameter files can be stored in an
ordinary writable folder. The temporary copy above is only needed
here because the example parameter file is part of the installed
package.

# One figure per parameter file

There is a second parameter file layout for cases where one parameter
file describes a single figure.

In the horizontal layout, each row is a figure and each column is a
parameter. In the vertical layout, each row is a parameter and the
two columns contain the parameter name and its value.

For example, the following describes a line chart:

```{r vertical}
params_vertical <- tempfile(fileext = ".csv")

writeLines(c(
  "plot_type,line",
  "title,Werkloosheid",
  "x,jaar",
  "y,werkloosheid",
  "ylab,%"
), params_vertical)

cat(readLines(params_vertical), sep = "\n")
```

The first column contains the parameter names:

```text
plot_type
title
x
y
ylab
```

The second column contains their values:

```text
line
Werkloosheid
jaar
werkloosheid
%
```

This layout can be convenient when a figure has many parameters,
because each setting has its own row.

When the parameter file describes one figure, `import_csv()` returns
that figure directly rather than a named list.

```{r vertical-build, results = "hide"}
fig_vertical <- import_csv(data_csv, params_vertical)
```

```{r vertical-build-show, echo = FALSE}
fig_vertical
```

# Handling problems

A parameter can be left blank. In that case, the default value is
used.

If a parameter is not valid for the selected `plot_type`, it is
ignored and the default is used. These cases produce a warning.

If a figure cannot be created, for example because a column
referenced by a parameter does not exist in the data file, that
figure is skipped. Other figures in the same parameter file are still
created.

Each call to `import_csv()` also writes a log file next to the
parameter file. The log records figures that were created or skipped
and any parameter problems.

# Run the import without opening R

The same CSV based method can also be used without opening R.

The `autogenerate_plots()` function creates a ready to use folder
containing:

* a data file
* a parameter file
* a script for Windows (a `.bat` file)
* a script for a Mac (a `.command` file)

After editing the data and parameter files, run the appropriate
script. It reads both files and saves the resulting figures as
images.

Running the script again after changing either CSV file produces
updated figures.

R and `ggcpb` must still be installed on the computer running the
script.

# Appendix: Parameter values {-}

The available parameters depend on the `plot_type`. The complete list
is stored in `parameter_settings.csv`.

```{r params-ref-path}
ref_csv <- system.file(
  "extdata",
  "parameter_settings.csv",
  package = "ggcpb"
)
```

The file gives the available parameters, their kind, default values
and examples.

```{r params-ref-table, echo = FALSE}
ref <- read.csv(
  ref_csv,
  stringsAsFactors = FALSE
)

knitr::kable(
  ref,
  col.names = c(
    "plot_type",
    "setting",
    "kind",
    "default",
    "example"
  )
)
```

A parameter of kind `column` takes a column name from the data file.
A parameter of kind `literal` takes a value directly from the
parameter file.

The `parameter_settings.csv` file can also be opened directly in a
spreadsheet program.
