---
title: "Usage in R: Loading data from CSV"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage in R: Loading data from CSV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Figures are drawn on the REAL CPB canvas (page = "half" is 2.98 in,
# "full" is 5.96 in; text sizes in theme_cpb() are absolute points, so
# any other canvas distorts the text-to-figure proportion) and only
# *displayed* smaller via out.width, at a consistent 2:1 scale:
# half-page figures at 350px, full-page figures at 700px.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
```

Most figures in this package are built by calling one of its
functions directly in R code. This document describes a different
way to build a figure. Two plain files hold everything that is
needed. One file holds the data. The other file lists the settings
for each figure. A single function reads both files and produces the
finished figures automatically. Very little knowledge of R is needed
to follow the steps below.

A separate option exists for building figures without opening R at
all. It is described briefly at the end of this document.

# The data file

Data that ships together with the package for demonstration purposes
is stored in a folder named `extdata`. This folder is part of the
package itself. The data file used in this document is named
`import_example_data.csv`.

The following code finds the exact location of that file on the
current computer, and stores it under the name `data_csv`. The exact
location is not shown here, since it depends on where the package
happens to be installed. Only the file name stays the same
everywhere.

```{r data-path}
data_csv <- system.file("extdata", "import_example_data.csv", package = "ggcpb")
```

The content of the file is shown below. This is exactly what would be
seen when opening the file directly, for example in a spreadsheet
program.

```{r data-raw, echo = FALSE, comment = ""}
cat(readLines(data_csv), sep = "\n")
```

Reading a CSV file into R turns it into a table. In R this kind of
table is called a data frame. Every row is one observation. Every
column is one variable.

```{r data-frame}
read.csv(data_csv)
```

The file has four columns. The column named `jaar` holds the year.
The column named `groep` holds the household group. The column named
`koopkracht` is the main value for each year and group. It stands for
the percentage change in purchasing power. The column named
`werkloosheid` is a second value. It stands for the unemployment
rate. This value repeats for every group within the same year,
because it is a single number for the whole year, not one number per
group. A repeated value like this is normal in a table of this
shape.

# The parameters file

A second file lists which figures to build, and how each one should
look. This file is called the parameters file. In this document it
is named `import_example_params.csv`.

```{r params-path}
params_csv <- system.file("extdata", "import_example_params.csv",
  package = "ggcpb")
```

```{r params-raw, echo = FALSE, comment = ""}
cat(readLines(params_csv), sep = "\n")
```

A parameters file can be written in one of two layouts. This example
uses the horizontal layout. Every row describes one figure. Every
column is one setting. The other layout is shown later in this
document.

Some settings refer to a column name from the data file. Examples are
`x`, `y`, `fill` and `sec_y`. Other settings are plain values, not
column names. Examples are `title`, `ylab` and `position`.

Two figures are described in this file.

The figure named `koopkracht` is a column chart. It shows
`koopkracht` by year. Bars for each group are placed side by side
rather than stacked on top of each other. Placing bars side by side
is called dodging. Without this setting the bars would be stacked.
Stacking would add the percentages of the three groups together. That
total does not have a clear meaning on its own. The `werkloosheid`
values are added to the same figure as a line, drawn against a second
axis on the right side.

The figure named `werkloosheid` is a plain line chart. It shows only
the `werkloosheid` values over time.

Every `plot_type` has its own list of settings. The full list,
together with which setting is a column reference and which is a
plain value, is stored in a reference file named
`parameter_settings.csv`, also inside the `extdata` folder. The same
information is available through the help page for the main
function, opened in R with a question mark followed by the function
name. The full table is also shown near the end of this document, in
the section named Parameter reference.

# Building the figures

One function call reads both files, and builds every figure listed in
the parameters file. Each resulting figure is stored under the name
given by its `id`.

```{r build, results = "hide"}
# The parameters file used here is part of the installed package and
# cannot be written to. A short log file is normally written next to
# the parameters file after every run, so a writable copy is made
# first, in a temporary folder. In everyday use the parameters file
# usually sits in an ordinary folder, so this extra step is normally
# not needed.
tmp_params <- tempfile(fileext = ".csv")
file.copy(params_csv, tmp_params, overwrite = TRUE)

figs <- import_csv(data_csv, tmp_params)
```

```{r fig-koopkracht, fig.width = 5.96, out.width = "700px"}
figs$koopkracht
```

```{r fig-werkloosheid}
figs$werkloosheid
```

Every call to the main function writes a plain text log file next to
the parameters file. The log lists what happened for each figure. It
states whether a figure was built or skipped, and gives the reason
for a skip. If a figure was built but one of its settings was not
recognised, that is listed as well. This record is useful when the
function runs as part of an automated process, where the console
output itself may not be seen by anyone.

The log for the two figures built above looks like this. The exact
file locations have been shortened to plain file names below, since
the full locations are specific to the computer this document was
built on:

```{r log, echo = FALSE, comment = ""}
log_path <- paste0(tools::file_path_sans_ext(tmp_params), "_log.txt")
log_text <- readLines(log_path)
log_text <- gsub(data_csv, "import_example_data.csv", log_text, fixed = TRUE)
log_text <- gsub(tmp_params, "import_example_params.csv", log_text,
  fixed = TRUE)
cat(log_text, sep = "\n")
```

# The vertical layout

Parameters can also be written in a second layout, called the
vertical layout. This layout suits a single figure with many settings
well. The first column lists the setting names, one per row. The
second column lists the values for that one figure.

```{r vertical}
params_vertical <- tempfile(fileext = ".csv")
writeLines(c(
  "plot_type,line",
  "title,Werkloosheid",
  "x,jaar",
  "y,werkloosheid",
  "ylab,%"
), params_vertical)
cat(readLines(params_vertical), sep = "\n")
```

When a parameters file describes only one figure, the main function
returns that figure directly, instead of a named list of figures.

```{r vertical-build}
import_csv(data_csv, params_vertical)
```

# When something goes wrong

A blank cell in the parameters file is simply skipped. The default
value for that setting is used instead. The same happens for a
setting name that does not belong to the chosen `plot_type`. In both
cases a warning is shown, and the same information is recorded in the
log file.

A figure that cannot be built at all is skipped in the same way, for
example when a setting refers to a column name that does not exist in
the data file. The reason is recorded in the log file. Other figures
in the same parameters file are still built normally. A problem in
one figure does not stop the rest from being produced.

# Parameter reference

The table below lists every setting for every `plot_type`, together
with its kind, its default, and an example value. A setting of kind
`column (from data_csv)` takes a column name from the data file. Its
example is a plausible column name, not a real one, since the actual
name always depends on the data file at hand. A setting of kind
`literal` takes a plain value instead. Its example is a value that
could be typed into the parameters file directly. A default shown as
several words separated by a forward slash lists every valid choice
for that setting, with the one used automatically named afterward.
An empty default means the setting is left out entirely unless it is
given a value.

The same table is stored as a plain CSV file, so it can also be
opened directly, for example in a spreadsheet program:

```{r params-ref-path}
ref_csv <- system.file("extdata", "parameter_settings.csv", package = "ggcpb")
```

```{r params-ref-table, echo = FALSE}
ref <- read.csv(ref_csv, stringsAsFactors = FALSE)
knitr::kable(ref, col.names = c("plot_type", "setting", "kind", "default", "example"))
```

# Building figures without opening R

A second option exists for the steps described above, meant for a
situation where R is not opened directly. The function
`cpb_import_kit()` copies a small, ready to use folder to a chosen
location. That folder contains an example data file and an example
parameters file, together with a script for Windows and a script for
a Mac. Opening the matching script with a double click reads both
files, and saves each figure as an image, without any further steps
in R. Editing the data file and the parameters file, then running the
script again, produces updated figures. R itself, and this package,
still need to be installed on the computer running the script.
