---
title: "Grouped layouts, facets & maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grouped layouts, facets & maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Figures are drawn on the REAL CPB canvas (page = "half" is 2.98 in,
# "full" is 5.96 in; text sizes in theme_cpb() are absolute points, so
# any other canvas distorts the text-to-figure proportion) and only
# *displayed* smaller via out.width, at a consistent 2:1 scale:
# half-page figures at 350px, full-page figures at 700px.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(42)
```

The basic chart types in `vignette("chart-types")` each show a single
chart. This vignette covers the ways CPB figures arrange *more than
one chart's worth* of information: a two-level category axis, the
vertically grouped distributional layout, small-multiple facets, and
choropleth maps of the Netherlands.

# Grouped categories on one shared axis

Pass `group` to `cpb_col()` to organise the categories into blocks --
a gap between the groups, the group names in bold under the category
labels, and *one* shared value axis (no facets). Each category must
belong to exactly one group; tune the gap with `group_gap`:

```{r col-grouped}
zakkans <- tibble(
  cat   = factor(rep(c("jongen", "meisje", "laag", "hoog"), each = 2),
                 levels = c("jongen", "meisje", "laag", "hoog")),
  grp   = factor(rep(c("geslacht", "opleiding ouders"), each = 4),
                 levels = c("geslacht", "opleiding ouders")),
  serie = rep(c("als centraal examen meetelt",
                "als centraal examen niet meetelt"), 4),
  pct   = c(8.7, 8.7, 9.7, 5.1, 11.2, 8.0, 8.1, 6.0))

cpb_col(zakkans, x = cat, y = pct, fill = serie, group = grp,
  position = "dodge", index = c(6, 2), width = 0.75,
  value_breaks = seq(0, 12, 2), value_limits = c(0, 12),
  title = "VWO", ylab = "zakkans (%)")
```

The group labels occupy the line an `xlab` would use, so the two
cannot be combined. This is the shared-axis alternative to faceting
(below): use it when the groups belong on one comparable scale.

# Vertically grouped boxplots

`cpb_box()` takes the same `group` argument for the published
distributional layout: categories organised under bold group
headings, all sharing one value axis. A group whose only category
carries the same name (like an "Alle huishoudens" total) collapses
onto its heading row. Combine with `box_style = "james"` (see
`vignette("chart-types")` for the box styles) and a *vector*
`fill_colour` -- one colour per row -- to colour each group:

```{r box-grouped, fig.height = 4.5}
ink <- tribble(
  ~cat,                     ~grp,                   ~p50,
  "Alle huishoudens",       "Alle huishoudens",     0.09,
  "1-20%",                  "Inkomensgroepen",      0.02,
  "21-40%",                 "Inkomensgroepen",      0.07,
  "41-60%",                 "Inkomensgroepen",      0.07,
  "61-80%",                 "Inkomensgroepen",      0.06,
  "81-100%",                "Inkomensgroepen",      0.04,
  "Werkenden",              "Inkomensbron",         0.09,
  "Uitkeringsgerechtigden", "Inkomensbron",         0.04,
  "Gepensioneerden",        "Inkomensbron",         0.05) |>
  mutate(cat = factor(cat, levels = cat),
         grp = factor(grp, levels = unique(grp)),
         p25 = p50 - runif(n(), 0.02, 0.15), p75 = p50 + runif(n(), 0.02, 0.12),
         p5  = p25 - runif(n(), 0.1, 0.6),   p95 = p75 + runif(n(), 0.1, 0.6))

cpb_box(ink, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  group = grp, box_style = "james", orientation = "horizontal",
  fill_colour = c("#193c69", rep("#87d2ff", 5), rep("#e6006e", 3)),
  width = 0.45,
  title = "Inkomenseffecten plannen stelsel",
  ylab  = "verandering in 2025 (%)")
```

The category rows keep the house ticks; the bold group headings carry
none and are outdented. A vertically grouped figure needs a taller
canvas -- here it is drawn 4.5 in tall.

# Facets

Every wrapper accepts a `facet` column. Facets follow the house
(legacy nicerplot) convention: the facet title is a bold strip *below*
each panel, and every panel is a complete mini-figure with its own
axes. Control the grid with `facet_ncol` and shared-versus-free axis
ranges with `facet_scales` (`"fixed"` by default, so panels are
directly comparable):

```{r facets, fig.width = 5.96, fig.height = 4.5, out.width = "700px"}
regios <- expand_grid(regio = factor(c("stad", "platteland", "gemengd", "totaal"),
                                     levels = c("stad", "platteland", "gemengd", "totaal")),
                      groep = factor(c("laag", "midden", "hoog"),
                                     levels = c("laag", "midden", "hoog")),
                      jaar  = 2019:2025) |>
  mutate(waarde = round(2 + as.numeric(groep) + cumsum(rnorm(n(), 0, 0.3)), 1))

cpb_col(regios, x = jaar, y = waarde, fill = groep,
  position   = "dodge",
  facet      = regio,
  facet_ncol = 2,
  index      = c(6, 2, 5),
  title = "Ontwikkeling per regio",
  ylab  = "mld euro")
```

A faceted figure usually needs a taller canvas: pass an explicit
`height` to `save_cpb()` (here the figure is drawn 4.5 in tall on the
full-page width). Facets versus `group`: reach for facets when each
panel is a self-contained chart, and for `group` when the categories
belong on one shared axis.

# Maps

`cpb_map()` draws a value per Dutch municipality, COROP region or
province on bundled generalised CBS/Kadaster boundaries (2025, via
cartomap). Regions are joined by CBS code (`"GM0014"`, `"PV20"`) or by
name, whichever matches best; regions without a value are filled with
the CPB missing-value grey. Borders are hairlines in the background
colour, so regions read as tiles separated by light-blue seams:

```{r map, fig.height = 3.4}
gemeenten <- tibble(code = unique(cpb_nl_geo("gemeente")$code)) |>
  mutate(index = rnorm(n(), 100, 15))

cpb_map(gemeenten, region = code, value = index,
  title    = "Voorbeeldindex per gemeente",
  subtitle = "index (Nederland = 100)")
```

Numeric values get the CPB sequential gradient; a discrete value
column gets the discrete palettes. Use `level = "corop"` or
`level = "provincie"` for the coarser boundaries. The raw boundary
tables are available through `cpb_nl_geo(level)` for anything the
wrapper does not cover.
