---
title: "Maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Figures are drawn on the REAL CPB canvas (page = "half" is 2.98 in,
# "full" is 5.96 in; text sizes in theme_cpb() are absolute points, so
# any other canvas distorts the text-to-figure proportion) and only
# *displayed* smaller via out.width, at a consistent 2:1 scale:
# half-page figures at 350px, full-page figures at 700px.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
library(ggplot2)
library(dplyr)
set.seed(42)
```

`cpb_map()` draws a value per Dutch municipality, COROP region or
province on bundled generalised CBS/Kadaster boundaries (2025, via
cartomap), so no geo packages or downloads are needed. Borders are
hairlines in the background colour, so regions read as tiles
separated by light-blue seams.

# Municipalities

Regions are joined by CBS code (`"GM0014"`, `"CR01"`, `"PV20"`) or by
name, whichever matches the `region` column best; regions without a
value are filled with the CPB missing-value grey, and unmatched
regions in the data raise a warning. A numeric `value` column gets
the CPB sequential gradient with a compact horizontal colourbar:

```{r map-gemeente, fig.height = 3.4}
gemeenten <- tibble(code = unique(cpb_nl_geo("gemeente")$code)) |>
  mutate(index = rnorm(n(), 100, 15))

cpb_map(gemeenten, region = code, value = index,
  title    = "Voorbeeldindex per gemeente",
  subtitle = "index (Nederland = 100)")
```

As elsewhere, the unit caption goes in `subtitle`; there is no value
axis, so `ylab` does not exist here.

# Provinces and COROP regions

`level` selects the boundaries -- `"provincie"` and `"corop"` for the
coarser levels -- and joining by *name* works too. A discrete `value`
column gets the discrete CPB palettes (pick colours with `index`):

```{r map-provincie, fig.height = 3.4}
provincies <- tibble(naam = unique(cpb_nl_geo("provincie")$name)) |>
  mutate(klasse = factor(
    sample(c("onder gemiddeld", "boven gemiddeld"), n(), replace = TRUE),
    levels = c("onder gemiddeld", "boven gemiddeld")
  ))

cpb_map(provincies, region = naam, value = klasse, level = "provincie",
  index = c(5, 6),
  title = "Groei ten opzichte van het landelijk gemiddelde")
```

# Styling and raw boundaries

The border seams are controlled with `border_colour` (default: the
CPB background colour -- a deliberate deviation from the legacy
plotter's outlined shapes) and `border_linewidth` (default `0.1`,
the house hairline). `reverse` flips the sequential gradient and
`na_fill` overrides the missing-value colour.

For anything the wrapper does not cover, the raw boundary tables are
available through `cpb_nl_geo(level)`: one row per polygon vertex
with the region `code` and `name`, ready for
`ggplot2::geom_polygon()` (use `part` as `group` and `ring` as
`subgroup`).

```{r geo}
head(cpb_nl_geo("provincie"), 3)
```
