---
title: "Maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Figures are drawn on the REAL CPB canvas (page = "half" is 2.98 in,
# "full" is 5.96 in; text sizes in theme_cpb() are absolute points, so
# any other canvas distorts the text-to-figure proportion) and only
# *displayed* smaller via out.width, at a consistent 2:1 scale:
# half-page figures at 350px, full-page figures at 700px.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "350px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
library(ggcpb)
library(ggplot2)
library(dplyr)
set.seed(42)
```

`cpb_map()` draws a value per Dutch municipality, COROP region or
province on bundled generalised CBS/Kadaster boundaries (2025, via
cartomap), so no geo packages or downloads are needed. Borders are
hairlines in the background colour, so regions read as tiles
separated by light-blue seams.

# Classed maps

CPB choropleths are usually *classed*: the values are binned into a
handful of ordered classes and filled with a light-to-dark ramp.
`cpb_cut()` does the binning with tidy Dutch class labels ("lager dan
20%", "20% - 30%", ..., "60% en hoger"), and `palette = "blues"` gives
the house blue ramp. Regions are joined by CBS code (`"GM0014"`,
`"CR01"`, `"PV20"`) or by name, whichever matches the `region` column
best; regions without a value get the CPB missing-value grey, and
unmatched regions raise a warning:

```{r map-classed, fig.height = 3.4}
gemeenten <- tibble(code = unique(cpb_nl_geo("gemeente")$code)) |>
  mutate(aandeel = runif(n(), 5, 75),
         klasse  = cpb_cut(aandeel, breaks = c(0, 20, 30, 40, 50, 60, Inf),
                           labeller = label_pct_nl()))

cpb_map(gemeenten, region = code, value = klasse,
  palette = "blues",
  title   = "Aandeel huishoudens met zonnepanelen",
  filllab = "aandeel")
```

`cpb_cut()` is a house-styled wrapper around `cut()`: give it the
`breaks` (including the outer bounds, `Inf` for an open top class) and
a formatter (`label_pct_nl()`, `label_euro_nl()`, `label_number_nl()`),
and it returns an ordered factor whose levels read the way published
figures label them. A single integer (`breaks = 5`) asks for that many
equal-width bins. The `"blues"` palette works in every CPB scale, not
just maps -- use it for classed bars too.

# Continuous maps

For a raw numeric value, `cpb_map()` fills with the CPB sequential
gradient and a compact horizontal colourbar. As elsewhere the unit
caption goes in `subtitle`; there is no value axis, so `ylab` does not
apply here:

```{r map-gemeente, fig.height = 3.4}
gemeenten_ct <- tibble(code = unique(cpb_nl_geo("gemeente")$code)) |>
  mutate(index = rnorm(n(), 100, 15))

cpb_map(gemeenten_ct, region = code, value = index,
  title    = "Voorbeeldindex per gemeente",
  subtitle = "index (Nederland = 100)")
```

# Provinces and COROP regions

`level` selects the boundaries -- `"provincie"` and `"corop"` for the
coarser levels -- and joining by *name* works too. A discrete `value`
column gets the discrete CPB palettes (pick colours with `index`):

```{r map-provincie, fig.height = 3.4}
provincies <- tibble(naam = unique(cpb_nl_geo("provincie")$name)) |>
  mutate(klasse = factor(
    sample(c("onder gemiddeld", "boven gemiddeld"), n(), replace = TRUE),
    levels = c("onder gemiddeld", "boven gemiddeld")
  ))

cpb_map(provincies, region = naam, value = klasse, level = "provincie",
  index = c(5, 6),
  title = "Groei ten opzichte van het landelijk gemiddelde")
```

# Styling and raw boundaries

The border seams are controlled with `border_colour` (default: the
CPB background colour -- a deliberate deviation from the legacy
plotter's outlined shapes) and `border_linewidth` (default `0.1`,
the house hairline). `reverse` flips the sequential gradient and
`na_fill` overrides the missing-value colour.

For anything the wrapper does not cover, the raw boundary tables are
available through `cpb_nl_geo(level)`: one row per polygon vertex
with the region `code` and `name`, ready for
`ggplot2::geom_polygon()` (use `part` as `group` and `ring` as
`subgroup`).

```{r geo}
head(cpb_nl_geo("provincie"), 3)
```
