---
title: "Text length stress test"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 2.98,
  fig.height = 2.98,
  dpi = 300,
  out.width = "450px",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
```

```{r setup, message = FALSE}
# knitr's working directory during a render is this file's own
# directory (inst/examples/), not the package root -- unlike an
# installed ggcpb, load_all() needs to be pointed at the root explicitly
devtools::load_all("../..", quiet = TRUE)
library(ggplot2)
library(dplyr)
library(tibble)
set.seed(1)
```

Not a curated example gallery -- a stress test. For every `cpb_*`
chart function, the same small dataset is drawn over and over with
only one thing changed each time: a short vs. long **title**,
**`ylab`**, **`xlab`**, or **tick label** (category names on the
category axis, or the magnitude of the values on the value axis).
`cpb_col()`, `cpb_box()` and `cpb_dot()` additionally get both
`orientation` options. Nothing here edits the package; it only calls
the existing functions with different inputs, so you can see where
long text clips, overlaps, or overflows the panel.

A title too long for the page is not wrapped automatically -- it just
clips at the panel edge, the same as any other ggplot2 plot. A manual
`"\n"` in `title` still works, since that is plain ggplot2 behaviour,
not something this package adds.

**The `ylab`/`xlab`/`subtitle` convention** (verified from the actual
wrapper code, not guessed): `ylab` always describes whichever axis
ends up drawn *vertically*, and -- left on its own, with no explicit
`subtitle` -- it is promoted into the subtitle line (an italic caption
above the panel) rather than a rotated axis title. `xlab` always
describes whichever axis ends up drawn *horizontally*, as an ordinary,
un-rotated axis title. Which one is "vertical" depends on
`orientation`: for `cpb_col()`/`cpb_box()` that is the value axis when
`orientation = "vertical"` and the category axis when `"horizontal"`;
`cpb_dot()`'s default (`"horizontal"`) already has categories
vertical. Passing an explicit `subtitle` *alongside* `ylab` is a
separate, documented edge case (the axis label then falls back to a
literal rotated title, or is dropped, depending on the function) --
tested here as its own case per function, not blended into the normal
usage.

```{r text-snippets}
title_short <- "Titel"
title_long  <- "Dit is een uitzonderlijk lange titel die het paneel ver overschrijdt en eigenlijk over twee regels opgesplitst zou moeten worden"

# ylab: normally promoted to the subtitle line
ylab_short <- "eenheid"
ylab_long  <- "Dit is een opvallend lange as-beschrijving die als onderschrift ver over de paneelbreedte heen loopt"

# xlab: an ordinary, un-rotated axis title
xlab_short <- "categorie"
xlab_long  <- "Een buitengewoon lange as-titel die veel horizontale ruimte in beslag neemt"

# a literal subtitle, for the "explicit subtitle + ylab together" edge case
subtitle_literal <- "(a) voorlopige cijfers, nog niet definitief vastgesteld"

cats_short <- c("Noord", "Oost", "Zuid", "West")
cats_long  <- c(
  "Een erg lange categorienaam die nooit helemaal past",
  "Nog een categorienaam die veel te lang is voor de as",
  "Deze categorienaam is ook onredelijk lang voor een tick label",
  "En de vierde lange categorienaam om het patroon door te zetten"
)
# even longer: full sentences, well past the ~18-character wrap width,
# to see how many lines a really extreme category name wraps onto
cats_xlong <- c(
  "Deze allereerste categorienaam is opzettelijk absurd lang gemaakt om te zien wat er gebeurt als een tick label vele regels beslaat",
  "De tweede categorienaam is minstens even lang en test of het paneel stabiel blijft ongeacht hoeveel tekst er in de as-labels staat",
  "Categorienaam nummer drie beschrijft in overdreven veel woorden een heel specifieke, gedetailleerde en langdradige beleidscategorie",
  "En de laatste, vierde categorienaam ronde het patroon af met een even onredelijk lange zin die de as tot het uiterste test"
)
```

# cpb_col()

`orientation = "vertical"` (default): the value axis is vertical, so
`ylab` names the value and lands on the subtitle line; `xlab` names
the category axis (usually left `NULL` -- the tick labels already
name the categories) as an ordinary bottom title.
`orientation = "horizontal"`: roles swap -- `xlab` now names the value
axis (drawn at the bottom, an ordinary title) and `ylab` names the
category axis (drawn on the left, promoted to the subtitle line) --
this is the real usage from the package's own examples.

```{r col-data}
col_short_dt <- tibble(cat = factor(cats_short, levels = cats_short),
                       waarde = c(12.4, 34.1, 8.7, 21.9))
col_long_dt  <- tibble(cat = factor(cats_long, levels = cats_long),
                       waarde = c(12.4, 34.1, 8.7, 21.9))
col_xlong_dt <- tibble(cat = factor(cats_xlong, levels = cats_xlong),
                       waarde = c(12.4, 34.1, 8.7, 21.9))
col_bignum_dt <- tibble(cat = factor(cats_short, levels = cats_short),
                        waarde = c(1234567, 3456789, 876543, 2198765))
```

## title only (no ylab/xlab)

```{r col-title-only}
cpb_col(col_short_dt, x = cat, y = waarde, title = title_short)
```

## long title

```{r col-long-title}
cpb_col(col_short_dt, x = cat, y = waarde, title = title_long)
```

## ylab alone -> subtitle (vertical), short then long

```{r col-ylab-subtitle-vertical}
cpb_col(col_short_dt, x = cat, y = waarde, title = title_short, ylab = ylab_short)
cpb_col(col_short_dt, x = cat, y = waarde, title = title_short, ylab = ylab_long)
```

## xlab alone -> ordinary title (vertical), short then long

```{r col-xlab-title-vertical}
cpb_col(col_short_dt, x = cat, y = waarde, title = title_short, xlab = xlab_short)
cpb_col(col_short_dt, x = cat, y = waarde, title = title_short, xlab = xlab_long)
```

## ylab + xlab together (vertical), short then long

```{r col-ylab-xlab-vertical}
cpb_col(col_short_dt, x = cat, y = waarde, title = title_short,
  ylab = ylab_short, xlab = xlab_short)
cpb_col(col_short_dt, x = cat, y = waarde, title = title_short,
  ylab = ylab_long, xlab = xlab_long)
```

## edge case: explicit subtitle + ylab together (vertical) -- documented fallback to a rotated axis title

```{r col-subtitle-plus-ylab-vertical}
cpb_col(col_short_dt, x = cat, y = waarde, title = title_short,
  subtitle = subtitle_literal, ylab = ylab_short)
```

## orientation = "horizontal": ylab -> category subtitle, xlab -> value title (the real usage)

```{r col-horizontal-real-usage}
cpb_col(col_short_dt, x = cat, y = waarde, orientation = "horizontal",
  title = title_short, ylab = xlab_short, xlab = ylab_short)
cpb_col(col_short_dt, x = cat, y = waarde, orientation = "horizontal",
  title = title_short, ylab = xlab_long, xlab = ylab_long)
```

## long x-axis tick labels (category names), vertical

```{r col-long-xticks, fig.height = 3.6}
cpb_col(col_long_dt, x = cat, y = waarde, title = title_short, ylab = ylab_short)
```

## long y-axis tick labels (large numbers), vertical

```{r col-long-yticks}
cpb_col(col_bignum_dt, x = cat, y = waarde, title = title_short, ylab = ylab_short)
```

## long category (tick) labels, horizontal

```{r col-long-cats-horizontal, fig.height = 3.2}
cpb_col(col_long_dt, x = cat, y = waarde, orientation = "horizontal",
  title = title_short, ylab = xlab_short)
```

## even longer category (tick) labels, horizontal -- known limit

Wrapping bounds the panel's *width* (each line is at most ~18
characters), but not row *height*: ggplot2's discrete axis does not
grow the spacing between categories to fit a taller, many-line wrapped
label, so full-sentence category names like these -- 7-10 wrapped
lines each -- collide into the row above/below instead. Realistic CPB
category names run at most to a short phrase (a sentence, at the
extreme), which is what wrapping targets; this is intentionally left
unhandled as a known limit rather than adding more machinery (e.g.
capping line count with an ellipsis) for text this far outside normal
use.

```{r col-xlong-cats-horizontal, fig.height = 3.2}
cpb_col(col_xlong_dt, x = cat, y = waarde, orientation = "horizontal",
  title = title_short, ylab = xlab_short)
```

# cpb_area()

No `orientation`: `ylab` always names the (vertical) value axis ->
subtitle; `xlab` is an ordinary bottom title for the x axis.

```{r area-data}
area_short_dt <- tibble(jaar = 2020:2024, waarde = c(10, 14, 9, 18, 22),
                        groep = factor("A"))
area_long_dt  <- tibble(jaar = 2020:2024, waarde = c(10, 14, 9, 18, 22),
                        groep = factor("Een erg lange legenda-categorienaam"))
```

## title only

```{r area-title-only}
cpb_area(area_short_dt, x = jaar, y = waarde, fill = groep, title = title_short)
```

## long title

```{r area-long-title}
cpb_area(area_short_dt, x = jaar, y = waarde, fill = groep, title = title_long)
```

## ylab alone -> subtitle, short then long

```{r area-ylab-subtitle}
cpb_area(area_short_dt, x = jaar, y = waarde, fill = groep, title = title_short, ylab = ylab_short)
cpb_area(area_short_dt, x = jaar, y = waarde, fill = groep, title = title_short, ylab = ylab_long)
```

## ylab + xlab together, long

```{r area-ylab-xlab-long}
cpb_area(area_short_dt, x = jaar, y = waarde, fill = groep, title = title_short,
  ylab = ylab_long, xlab = xlab_long)
```

## edge case: explicit subtitle + ylab together -- fallback to a rotated axis title

```{r area-subtitle-plus-ylab}
cpb_area(area_short_dt, x = jaar, y = waarde, fill = groep, title = title_short,
  subtitle = subtitle_literal, ylab = ylab_short)
```

## long y-axis tick labels (large numbers)

```{r area-long-yticks}
area_bignum_dt <- area_short_dt |> mutate(waarde = waarde * 1e6 + 123456)
cpb_area(area_bignum_dt, x = jaar, y = waarde, fill = groep, title = title_short, ylab = ylab_short)
```

## long legend label (fill category)

```{r area-long-legend}
cpb_area(area_long_dt, x = jaar, y = waarde, fill = groep, title = title_short, ylab = ylab_short)
```

# cpb_line()

Same convention as `cpb_area()` (no `orientation`).

```{r line-data}
line_short_dt <- tibble(jaar = 2015:2024,
                        waarde = 100 + cumsum(rnorm(10, 1, 1.2)))
```

## title only

```{r line-title-only}
cpb_line(line_short_dt, x = jaar, y = waarde, title = title_short)
```

## long title

```{r line-long-title}
cpb_line(line_short_dt, x = jaar, y = waarde, title = title_long)
```

## ylab alone -> subtitle, short then long

```{r line-ylab-subtitle}
cpb_line(line_short_dt, x = jaar, y = waarde, title = title_short, ylab = ylab_short)
cpb_line(line_short_dt, x = jaar, y = waarde, title = title_short, ylab = ylab_long)
```

## ylab + xlab together, long

```{r line-ylab-xlab-long}
cpb_line(line_short_dt, x = jaar, y = waarde, title = title_short,
  ylab = ylab_long, xlab = xlab_long)
```

## edge case: explicit subtitle + ylab together -- fallback to a rotated axis title

```{r line-subtitle-plus-ylab}
cpb_line(line_short_dt, x = jaar, y = waarde, title = title_short,
  subtitle = subtitle_literal, ylab = ylab_short)
```

## long y-axis tick labels (large numbers)

```{r line-long-yticks}
line_bignum_dt <- line_short_dt |> mutate(waarde = waarde * 1e5 + 123456)
cpb_line(line_bignum_dt, x = jaar, y = waarde, title = title_short, ylab = ylab_short)
```

# cpb_box()

Same `orientation`-dependent role swap as `cpb_col()`.

```{r box-data}
box_short_dt <- tibble(cat = factor(cats_short, levels = cats_short),
                       p5 = c(-2, -1, -3, 0), p25 = c(0, 1, -1, 2),
                       p50 = c(1, 2.5, 0, 3.5), p75 = c(2, 4, 1, 5),
                       p95 = c(4, 6, 3, 7))
box_long_dt <- tibble(cat = factor(cats_long, levels = cats_long),
                      p5 = c(-2, -1, -3, 0), p25 = c(0, 1, -1, 2),
                      p50 = c(1, 2.5, 0, 3.5), p75 = c(2, 4, 1, 5),
                      p95 = c(4, 6, 3, 7))
box_bignum_dt <- box_short_dt |> mutate(across(p5:p95, ~ .x * 1e5 + 123456))
```

## title only, vertical vs. horizontal

```{r box-title-only}
cpb_box(box_short_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "vertical", title = title_short)
cpb_box(box_short_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "horizontal", title = title_short)
```

## long title, vertical vs. horizontal

```{r box-long-title}
cpb_box(box_short_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "vertical", title = title_long)
cpb_box(box_short_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "horizontal", title = title_long)
```

## ylab -> value subtitle (vertical), short then long

```{r box-ylab-subtitle-vertical}
cpb_box(box_short_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "vertical", title = title_short, ylab = ylab_short)
cpb_box(box_short_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "vertical", title = title_short, ylab = ylab_long)
```

## ylab -> category subtitle (horizontal), short then long

```{r box-ylab-subtitle-horizontal}
cpb_box(box_short_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "horizontal", title = title_short, ylab = xlab_short)
cpb_box(box_short_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "horizontal", title = title_short, ylab = xlab_long)
```

## xlab -> value title (horizontal), short then long

```{r box-xlab-title-horizontal}
cpb_box(box_short_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "horizontal", title = title_short, xlab = ylab_short)
cpb_box(box_short_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "horizontal", title = title_short, xlab = ylab_long)
```

## edge case: explicit subtitle + ylab together, vertical -- documented fallback

```{r box-subtitle-plus-ylab-vertical}
cpb_box(box_short_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "vertical", title = title_short,
  subtitle = subtitle_literal, ylab = ylab_short)
```

## long category (tick) labels, vertical vs. horizontal

```{r box-long-cats, fig.height = 3.8}
cpb_box(box_long_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "vertical", title = title_short, ylab = ylab_short)
```

```{r box-long-cats-horizontal, fig.height = 3.2}
cpb_box(box_long_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "horizontal", title = title_short, ylab = xlab_short)
```

## long value-axis tick labels (large numbers), vertical

```{r box-long-yticks-vertical}
cpb_box(box_bignum_dt, x = cat, p5 = p5, p25 = p25, p50 = p50, p75 = p75, p95 = p95,
  orientation = "vertical", title = title_short, ylab = ylab_short)
```

# cpb_scatter()

No `orientation`: `ylab` -> subtitle, `xlab` -> ordinary bottom title
(this one is `xlab`-relevant, unlike the category-driven wrappers,
since x is a genuine numeric axis here).

```{r scatter-data}
scatter_dt <- tibble(x = rnorm(30, 10, 3), y = rnorm(30, 5, 2))
```

## title only

```{r scatter-title-only}
cpb_scatter(scatter_dt, x = x, y = y, title = title_short)
```

## long title

```{r scatter-long-title}
cpb_scatter(scatter_dt, x = x, y = y, title = title_long)
```

## ylab + xlab together, short then long

```{r scatter-ylab-xlab}
cpb_scatter(scatter_dt, x = x, y = y, title = title_short, ylab = ylab_short, xlab = xlab_short)
cpb_scatter(scatter_dt, x = x, y = y, title = title_short, ylab = ylab_long, xlab = xlab_long)
```

## edge case: explicit subtitle + ylab together -- fallback to a rotated axis title

```{r scatter-subtitle-plus-ylab}
cpb_scatter(scatter_dt, x = x, y = y, title = title_short,
  subtitle = subtitle_literal, ylab = ylab_short)
```

## long axis tick labels (large numbers, both axes)

```{r scatter-long-ticks}
scatter_bignum_dt <- scatter_dt |> mutate(x = x * 1e6 + 123456, y = y * 1e6 + 654321)
cpb_scatter(scatter_bignum_dt, x = x, y = y, title = title_short, ylab = ylab_short, xlab = xlab_short)
```

# cpb_hist()

Same convention as `cpb_scatter()`.

```{r hist-data}
hist_dt <- tibble(x = rnorm(300, 100, 15))
hist_bignum_dt <- tibble(x = rnorm(300, 100, 15) * 1e5 + 123456)
```

## title only

```{r hist-title-only}
cpb_hist(hist_dt, x = x, title = title_short)
```

## long title

```{r hist-long-title}
cpb_hist(hist_dt, x = x, title = title_long)
```

## ylab + xlab together, short then long

```{r hist-ylab-xlab}
cpb_hist(hist_dt, x = x, title = title_short, ylab = ylab_short, xlab = xlab_short)
cpb_hist(hist_dt, x = x, title = title_short, ylab = ylab_long, xlab = xlab_long)
```

## edge case: explicit subtitle + ylab together -- fallback to a rotated axis title

```{r hist-subtitle-plus-ylab}
cpb_hist(hist_dt, x = x, title = title_short,
  subtitle = subtitle_literal, ylab = ylab_short)
```

## long x-axis tick labels (large numbers)

```{r hist-long-xticks}
cpb_hist(hist_bignum_dt, x = x, title = title_short, ylab = ylab_short, xlab = xlab_short)
```

# cpb_dot()

Default `orientation = "horizontal"`: categories are already vertical
(left), values already horizontal (bottom) -- `ylab` names the
category axis (-> subtitle), `xlab` names the value axis (-> ordinary
bottom title). `orientation = "vertical"` flips both roles.

```{r dot-data}
dot_short_dt <- tibble(cat = factor(cats_short, levels = cats_short),
                       y = c(1, 2.5, 0, 3.5), lower = c(0, 1, -1, 2), upper = c(2, 4, 1, 5))
dot_long_dt <- tibble(cat = factor(cats_long, levels = cats_long),
                      y = c(1, 2.5, 0, 3.5), lower = c(0, 1, -1, 2), upper = c(2, 4, 1, 5))
dot_bignum_dt <- dot_short_dt |> mutate(across(c(y, lower, upper), ~ .x * 1e5 + 123456))
```

## title only, horizontal vs. vertical

```{r dot-title-only}
cpb_dot(dot_short_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "horizontal", title = title_short)
cpb_dot(dot_short_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "vertical", title = title_short)
```

## long title, horizontal vs. vertical

```{r dot-long-title}
cpb_dot(dot_short_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "horizontal", title = title_long)
cpb_dot(dot_short_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "vertical", title = title_long)
```

## ylab -> category subtitle, xlab -> value title (horizontal), short then long

```{r dot-ylab-xlab-horizontal}
cpb_dot(dot_short_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "horizontal", title = title_short, ylab = xlab_short, xlab = ylab_short)
cpb_dot(dot_short_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "horizontal", title = title_short, ylab = xlab_long, xlab = ylab_long)
```

## ylab -> value subtitle (vertical), short then long

```{r dot-ylab-subtitle-vertical}
cpb_dot(dot_short_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "vertical", title = title_short, xlab = ylab_short)
cpb_dot(dot_short_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "vertical", title = title_short, xlab = ylab_long)
```

## long category (tick) labels, horizontal vs. vertical

```{r dot-long-cats, fig.height = 3.2}
cpb_dot(dot_long_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "horizontal", title = title_short, ylab = xlab_short)
```

```{r dot-long-cats-vertical, fig.height = 3.8}
cpb_dot(dot_long_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "vertical", title = title_short, xlab = ylab_short)
```

## manual "\n" in the category values themselves (not auto-wrap)

A category label can also be split by hand -- a literal `"\n"` typed
into the data (`"Noord\nregio"`), not the automatic ~18-character
wrap above -- and that manual break is kept exactly as given, the same
way a manual `"\n"` in a title is (see "Title wrapping"): only an
individual line that is itself still too wide for the axis gets
wrapped further, on its own.

```{r dot-manual-cats, fig.height = 3.2}
dot_manual_dt <- dot_short_dt |>
  mutate(cat = factor(paste0(as.character(cat), "\nregio"),
                      levels = paste0(as.character(cat), "\nregio")))
cpb_dot(dot_manual_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "horizontal", title = title_short, ylab = xlab_short)
```

## long value-axis tick labels (large numbers), horizontal

```{r dot-long-yticks-horizontal}
cpb_dot(dot_bignum_dt, x = cat, y = y, lower = lower, upper = upper,
  orientation = "horizontal", title = title_short, xlab = ylab_short)
```

# cpb_donut()

No `ylab`/`xlab` here -- there is no axis to label. Printed directly
(not through `save_cpb()`): `cpb_donut()`'s `panel_size` (a fixed
physical ring size regardless of chrome) does not apply outside
`save_cpb()`, so the ring itself still reflows with the surrounding
chrome here, same as any other wrapper's panel.

```{r donut-data}
donut_short_dt <- tibble(cat = factor(cats_short, levels = cats_short),
                         waarde = c(35, 30, 20, 15))
donut_long_dt <- tibble(cat = factor(cats_long, levels = cats_long),
                        waarde = c(35, 30, 20, 15))
```

## baseline (everything short)

```{r donut-baseline}
cpb_donut(donut_short_dt, fill = cat, y = waarde,
  title = title_short, subtitle = ylab_short)
```

## long title + subtitle

```{r donut-long-title}
cpb_donut(donut_short_dt, fill = cat, y = waarde,
  title = title_long, subtitle = ylab_long)
```

## long legend labels (fill categories)

```{r donut-long-legend}
cpb_donut(donut_long_dt, fill = cat, y = waarde,
  title = title_short, subtitle = ylab_short)
```

## long wedge labels, label_style = "wedge" (wraps)

```{r donut-long-wedge-labels-wedge}
cpb_donut(donut_long_dt, fill = cat, y = waarde, label = cat,
  label_style = "wedge",
  title = title_short, subtitle = ylab_short)
```

## long wedge labels, label_style = "leader" (wraps too, capped at 3 lines)

`"leader"` places its labels with its own collision-avoidance math
(spacing same-side labels apart vertically) -- capping the wrap at 3
lines keeps that math's assumed label height bounded, however long
the input text runs, rather than leaving it unwrapped and unbounded.

```{r donut-long-wedge-labels-leader}
cpb_donut(donut_long_dt, fill = cat, y = waarde, label = cat,
  label_style = "leader",
  title = title_short, subtitle = ylab_short)
```

# cpb_map()

`cpb_map()` has no conventional x/y axes (the panel is the
choropleth itself), so only title/subtitle/legend length are relevant
here, not tick labels or `ylab`/`xlab`. Every map goes through
`save_cpb()`, the same way `vignette("maps")` does, since that is what
auto-fits the panel to the boundaries' true aspect ratio.

```{r show_map}
show_map <- function(...) {
  f <- tempfile(fileext = ".png")
  save_cpb(f, cpb_map(...), page = "half")
  knitr::include_graphics(f)
}
```

```{r map-data}
map_dt <- tibble(naam = unique(cpb_nl_geo("provincie")$name),
                 waarde = rnorm(12, 100, 12))
```

## baseline (everything short)

```{r map-baseline}
show_map(map_dt, region = naam, value = waarde, level = "provincie",
  title = title_short, subtitle = ylab_short)
```

## long title + subtitle

```{r map-long-title}
show_map(map_dt, region = naam, value = waarde, level = "provincie",
  title = title_long, subtitle = ylab_long)
```

## long legend title (filllab)

```{r map-long-filllab}
show_map(map_dt, region = naam, value = waarde, level = "provincie",
  title = title_short, subtitle = ylab_short, filllab = xlab_long)
```
